# Regression

## Working Example {data-toc-title="Example"}

Let's assume a SQUARE test exists

- with test results between 0 and 20
- performed on patients
- who have particular characteristics (e.g., age, ...)
- under particular conditions (e.g., treated or not)

Aim: understand the conditions (e.g., treatment) under which test results are higher/lower, accounting for age

- by relating data (test results ~ outcome) to conditions and characteristics (predictors) &rarr; columns
- using information offered by each of the patients (replications ~ sample size) &rarr; rows

Equation-wise:

$$
square\ test\ result_{id} = \beta_0 + \beta_1 * Treatment_{id} + \beta_2 * Age_{id} + \varepsilon_{id},
\qquad \text{for } id = 1,\dots,n,
$$

More generically:

$$
y_i = \beta_0 + \beta_1 x1_i + \beta_2 x2_i + \varepsilon_i,
\qquad \text{for } i = 1,\dots,n.
$$

Regression (and ANOVA)

- is modeling variance (e.g., in $y_i$)
- turning it into a conditional mean (e.g., in $\beta_0 + \beta_1 * Treatment_{id} + \beta_2 * Age_{id}$)
- by minimizing squared error (e.g., of $\varepsilon_i^2$).

## Regression Model: Mean and Variance {data-toc-title="Mean - Variance"}

Regression as swiss knife.

:::: {.columns}

A regression model with only an intercept extracts 

- mean (conditional on nothing else)
- variance (of the residuals around the mean)  
- standard error (uncertainty around the estimated mean)

$$ y_i = \beta_0 + \varepsilon_i \qquad with \qquad \varepsilon_i \overset{\text{i.i.d.}}{\sim} \mathcal{N}(0, \sigma^2). $$

&nbsp;

::: {.column width="50%" .light-grey-col}

```{r}
#| label: fig-intercept_only
#| fig-cap: "Intercept-only model"
#| fig-height: 5.5
#| fig-width: 7

ybar <- mean(dta_base$score)

s <- sd(scores)
se <- s / sqrt(n)

alpha_ci <- 0.05
t_crit <- qt(1 - alpha_ci/2, df = n - 1)
ci_lower <- ybar - t_crit * se
ci_upper <- ybar + t_crit * se

ggplot(dta_base, aes(x = id, y = score)) +
  geom_rect(
    xmin = 0.5,
    xmax = n + 0.5,
    ymin = mu_true - sigma,
    ymax = mu_true + sigma,
    fill = "gray80",
    alpha = 0.3,
    inherit.aes = FALSE
  ) +
  geom_rect(
    xmin = 0.5,
    xmax = n + 0.5,
    ymin = ci_lower,
    ymax = ci_upper,
    fill = "gray60",
    alpha = 0.25,
    inherit.aes = FALSE
  ) +
  annotate(
    "rect",
    xmin = n + 1.5,
    xmax = n + 2,
    ymin = ybar - se,
    ymax = ybar + se,
    fill = "#f60",
    alpha = 0.8
  ) +
  geom_point(
    size = 2,
    color = "#333333",
    alpha = 0.9
  ) +
  geom_segment(
    aes(x = id, xend = id, y = score, yend = ybar),
    color = "#666666",
    alpha = 0.4,
    linewidth = 1
  ) +
  geom_hline(
    yintercept = ybar,
    color = "#1f78b4",
    linewidth = 2,
    linetype = "dashed"
  ) +
  annotate(
    "text",
    x = n - 1,
    y = ybar + 0.5,
    label = "SE",
    color = "#f60",
    hjust = -1.1,
    size = 5
  ) +
  annotate(
    "text",
    x = 2,
    y = ybar + 0.8,
    label = paste0("Intercept = mean ≈ ", round(ybar, 2)),
    color = "#039",
    hjust = 0,
    size = 5
  ) +
  annotate(
    "segment",
    x = 3,
    xend = 4,
    y = mu_true + sigma,
    yend = mu_true + sigma,
    color = "gray70",
    linewidth = 2,
    alpha = 0.6
  ) +
  annotate(
    "text",
    x = 5,
    y = mu_true + sigma,
    label = "±1 SD",
    hjust = 0,
    color = "gray30",
    size = 3.5
  ) +
  labs(
    x = "Patient",
    y = "Result",
    title = "Intercept-only model"
  ) + 
  coord_cartesian(ylim = c(10, 20)) 
```

::: {.smaller}
Patients on x-axis, observations on y-axis

* line represents average
* greyed band +/- 1 standard deviation
:::

:::

::: {.column width="5%"}

:::

::: {.column width="45%"}

Fitted intercept is the mean,  
or average SQUARE test result

$$
\hat{\beta}_0 = \bar{y}
$$

Residuals are difference data - model based expectation.

$$
e_i = y_i - \bar{y}
$$

No variance explained, with total variance $\sigma^2$.

$$
\sigma^2 = \frac{1}{N} \sum_{i=1}^{N} (y_i - \mu)^2
$$

Predicts the same value for every case.

$$
SE(\hat{\beta}_0) = \frac{\sigma}{\sqrt{n}}
$$

The standard error (SE; estimation) is a function of the standard deviation (sd; data). 
:::


```{r}
#| label: lm_intercept
#| echo: true
lm_b0 <- lm(scores ~ 1, data = dta_base)
```

```{r}
#| label: tab-tidy_lm_intercept_only
#| tab-cap: "Result intercept only"
tidy_gt(lm_b0)
glance_gt(lm_b0)
```
::::


## Two Groups; t-test look-alike {data-toc-title="Two groups"}

:::: {.columns}

A regression model with an intercept and a treatment effect (slope, dichotomous variable Treatment vs. Control)

- treatment and control specific means
- difference between treatment and control
- standard error of difference

The regression model is:

$$
y_i = \beta_0 + \underline{\beta_1 x_i} + \varepsilon_i,
\qquad \text{for } i = 1,\dots,n,
$$

With

$$
x_i =
\begin{cases}
0 & \text{for control}, \\
1 & \text{for treatment}.
\end{cases}
$$

And the usual distributional assumptions.

$$
\varepsilon_i \overset{\text{i.i.d.}}{\sim} \mathcal{N}(0, \sigma^2).
$$

&nbsp;

::: {.column width="50%" .light-grey-col}

```{r}
#| label: fig-two_groups
#| fig-cap: "Two group-specific means"
#| fig-height: 5.5
#| fig-width: 7
# use dta_groups, col1, col2 
grp_stats <- dta_groups |>
  group_by(group) |>
  summarise(
    mean = mean(score),
    sd = sd(score),
    n = n(),
    se = sd(score) / sqrt(n()),
    .groups = "drop"
  ) |>
  mutate(
    alpha_ci = 0.05,
    t_crit   = qt(1 - alpha_ci/2, df = n - 1),
    ci_lower = mean - t_crit * se,
    ci_upper = mean + t_crit * se
  )

pltdta_groups <- dta_groups |>
  left_join(grp_stats |> select(group, mean), by = "group")

ggplot(pltdta_groups, aes(x = id, y = score, color = group, fill = group)) +
  geom_point(size = 2, alpha = 0.9) +
  scale_color_manual(values = c("control" = col1, "treatment" = col2)) +
  geom_segment(
    data = pltdta_groups,
    aes(x = id, xend = id, y = score, yend = mean),
    color = "#666666",
    alpha = 0.4,
    linewidth = 1
  ) +
  geom_hline(
    data = grp_stats,
    aes(yintercept = mean, color = group),
    linewidth = 2.5,
    linetype = "dashed"
  ) +
  geom_text(
    data = data.frame(
      x = 1.2,
      y = grp_stats$mean[1] + 0.7,
      label = paste0("control mean ≈ ", round(grp_stats$mean[1], 2)),
      group = "Control"
    ),
    aes(x = x, y = y, label = label, color = group),
    hjust = 0,
    size = 4,
    inherit.aes = FALSE
  ) +
  geom_text(
    data = data.frame(
      x = 1.2,
      y = grp_stats$mean[2] + 0.7,
      label = paste0("treatment mean ≈ ", round(grp_stats$mean[2], 2)),
      group = "Treatment"
    ),
    aes(x = x, y = y, label = label, color = group),
    hjust = 0,
    size = 4,
    inherit.aes = FALSE
  ) +
  geom_rect(
    data = data.frame(
      xmin = c(0.5, n1 + 0.75),
      xmax = c(n1 + 0.25, n + 0.5),
      ymin = grp_stats$mean - grp_stats$sd,
      ymax = grp_stats$mean + grp_stats$sd,
      group_fill = c(col1, col2)
    ),
    aes(xmin = xmin, xmax = xmax, ymin = ymin, ymax = ymax, fill = group_fill),
    alpha = 0.3,
    inherit.aes = FALSE
  ) +
   scale_fill_manual(values = c("control" = col1, "treatment" = col2), guide = "none") +
  annotate(
    "text",
    x = 8,
    y = grp_stats$mean[1] - grp_stats$sd[1],
    label = "±1 SD (G1)",
    hjust = 0,
    color = "gray30",
    size = 3.5
  ) +
  annotate(
    "text",
    x = 40,
    y = grp_stats$mean[2] + grp_stats$sd[2],
    label = "±1 SD (G2)",
    hjust = 0,
    color = "gray30",
    size = 3.5
  ) +
  labs(x = "Patient", y = "Result", title = "Two-group model", color="") + 
  coord_cartesian(ylim = c(10, 20)) +
  guides(color = guide_legend(override.aes = list(size = 3))) 
```

::: {.smaller}
Patients on x-axis, observations on y-axis

* line represents group specific means
* greyed band +/- 1 group specific standard deviation
:::

:::

::: {.column width="5%"}

:::

::: {.column width="45%"}

Fitted intercept is the mean for group coded as 0,  
fitted slope is the difference between the two groups
offer estimated group means (conditional means)

$$
\hat{\beta}_0 = \bar{y}_{x=0}
$$
$$
\hat{\beta}_1 = \bar{y}_{x=1} - \bar{y}_{x=0}
$$

Residuals are still difference data - model based expectation.

$$
e_i = y_i - \hat{y}_i
$$

Notice: residuals are defined over all observations, not per condition.

<!--
$$
SE(\hat{\beta}_1) = \sqrt{\frac{sd_0^2}{n_0} + \frac{sd_1^2}{n_1}}
$$
-->

The standard error (SE; estimation) is a function of the group specific standard deviations (sd; data) and sample sizes. 

About `r round(summary(lm_groups)$r.squared,4)*100` % of the variance is now explained.

Predicts the same value for every case within a group.
:::


```{r}
#| label: lm_groups
#| echo: true
lm_groups <- lm(score ~ group, data = dta_groups)
```

```{r}
#| label: tab-tidy_lm_groups
#| tab-cap: "Result 2 groups"
tidy_gt(lm_groups)
glance_gt(lm_groups)
```

::::

## Continuous Variable {data-toc-title="Continuous"}

:::: {.columns}

A regression model with an intercept and a continuous predictor (slope)

- predictor specific means
- difference for 1 unit extra on the predictor
- standard error of the slope

The regression model is exactly the same as before:

$$
y_i = \beta_0 + \underline{\beta_1 x_i} + \varepsilon_i,
\qquad \text{for } i = 1,\dots,n,
$$

with the usual distributional assumption:

$$
\varepsilon_i \overset{\text{i.i.d.}}{\sim} \mathcal{N}(0, \sigma^2).
$$

&nbsp;

::: {.column width="50%" .light-grey-col}

```{r}
#| label: fig-continuous_reordered
#| fig-cap: "Predictor value specific means"
#| fig-height: 5.5
#| fig-width: 7
# use dta_cont
lm_cont <- lm(score ~ age, data = dta_cont)

cont_stats <- dta_cont |>
  summarise(
    mean_score = mean(score),
    sd_score   = sd(score),
    n          = n(),
    se_mean    = sd(score) / sqrt(n()),
    .groups    = "drop"
  )

pltdta_cont <- dta_cont |>
  mutate(
    fitted = predict(lm_cont),
    resid  = score - fitted
  )

col_cont <- col1

ggplot(pltdta_cont, aes(x = age, y = score)) +
  geom_point(size = 2, alpha = 0.9, color = col_cont) +
  geom_segment(
    data = pltdta_cont,
    aes(x = age, xend = age, y = score, yend = fitted),
    color = "gray60",
    alpha = 0.4,
    linewidth = 1
  ) +
  geom_smooth(
    method = "lm",
    se = TRUE,
    color = col_cont,
    linewidth = 1.5,
    alpha = 0.2,
    fullrange = TRUE
  ) +
  geom_hline(
    data = cont_stats,
    aes(yintercept = mean_score),
    color = "gray60",
    linewidth = 2,
    linetype = "dashed",
    alpha = 0.8
  ) +
  geom_rect(
    data = data.frame(
      xmin = max(pltdta_cont$age) + 0.5,
      xmax = max(pltdta_cont$age) + 1,
      ymin = cont_stats$mean_score - cont_stats$se_mean,
      ymax = cont_stats$mean_score + cont_stats$se_mean
    ),
    aes(xmin = xmin, xmax = xmax, ymin = ymin, ymax = ymax),
    fill = col2,
    alpha = 0.4,
    color = "black",
    linewidth = 0.5,
    inherit.aes = FALSE
  ) +
  geom_rect(
    data = data.frame(
      xmin = min(pltdta_cont$age) - 1,
      xmax = max(pltdta_cont$age) + 1,
      ymin = cont_stats$mean_score - cont_stats$sd_score,
      ymax = cont_stats$mean_score + cont_stats$sd_score
    ),
    aes(xmin = xmin, xmax = xmax, ymin = ymin, ymax = ymax),
    fill = "gray70",
    alpha = 0.15,
    inherit.aes = FALSE
  ) +
  annotate(
    "text",
    x = max(pltdta_cont$age) - 1,
    y = cont_stats$mean_score + 0.3,
    label = "SE",
    color = col2,
    hjust = -0.1,
    size = 5
  ) +
  annotate(
    "text",
    x = min(pltdta_cont$age) + 1,
    y = cont_stats$mean_score + cont_stats$sd_score,
    label = "±1 SD",
    hjust = 0,
    color = "gray30",
    size = 3.5
  ) +
  annotate(
    "text",
    x = mean(pltdta_cont$age),
    y = max(pltdta_cont$score) - 0.8,
    label = paste0(
      "Linear trend: score ≈ ",
      round(coef(lm_cont)[1], 2),
      " + ",
      round(coef(lm_cont)[2], 3),
      " × age"
    ),
    size = 4.5
  ) +
  labs(
    x = "Age (continuous predictor)",
    y = "Score (outcome)",
    title = "Continuous predictor model"
  ) + 
  coord_cartesian(ylim = c(10, 20)) 
```

::: {.smaller}
Note, the same 64 scores are used, but treated as continuous predictor $x_i$.

Predictor $x_i$ on **x-axis** (age), observations on y-axis

* line represents **linear trend**
* greyed band +/- 1 **overall** standard deviation
:::

:::

::: {.column width="5%"}

:::

::: {.column width="45%"}

Fitted intercept is the mean at $x_i$ = 0  
Fitted slope is increase per unit $x_i$,  

Note: center your $x_i$ around another value,  
change the intercept.

$$
\hat{\beta}_0 = \mathbb{E}[Y \mid X=0]
$$

$$
\hat{\beta}_1 = \frac{\Delta \mathbb{E}[Y]}{\Delta X}
$$

Residuals are difference  
data - model based expectation.

$$
e_i = y_i - \hat{y}_i
$$

$$
SE(\hat{\beta}_1) = \sqrt{\frac{sd^2}{n \cdot \text{Var}(X)}}
$$

The standard error (SE; estimation) is a function of the **residual variance** relative to predictor spread.
:::

```{r}
#| label: lm_cont
#| echo: true
lm_cont <- lm(score ~ age, data = dta_cont)
```

```{r}
#| label: tab-tidy_lm_cont
#| tab-cap: "Result continuous predictor"
tidy_gt(lm_cont)
glance_gt(lm_cont) 
```
::::


## Dichotomized continuous predictor {data-toc-title="Dichotomized"}

Same model !

:::: {.columns}

A regression model with an intercept and a continuous predictor that is dichotomized into two groups.

&nbsp;

::: {.column width="50%" .light-grey-col}

```{r}
#| label: fig-continuous_dichotomized
#| fig-cap: "Dichotomized age groups"
#| fig-height: 5.5
#| fig-width: 7
# use dta_dich, col1, col2

grp_stats <- dta_dich |>
  group_by(age_d) |>
  summarise(
    mean = mean(score),
    sd = sd(score),
    n = n(),
    se = sd(score) / sqrt(n()),
    .groups = "drop"
  ) |>
  mutate(
    alpha_ci = 0.05,
    t_crit   = qt(1 - alpha_ci/2, df = n - 1),
    ci_lower = mean - t_crit * se,
    ci_upper = mean + t_crit * se
  )

n1 <- sum(dta_dich$age_d == "age ≤60")
n2 <- sum(dta_dich$age_d == "age >60")

pltdta_dich <- dta_dich |>
  left_join(grp_stats |> select(age_d, mean), by = "age_d")

ggplot(pltdta_dich, aes(x = age, y = score, color = age_d, fill = age_d)) +
  geom_point(size = 2, alpha = 0.9) +
  scale_color_manual(values = c("age ≤60" = col1, "age >60" = col2)) +
  geom_segment(
    data = pltdta_dich,
    aes(x = age, xend = age, y = score, yend = mean),
    color = "#666666",
    alpha = 0.4,
    linewidth = 1
  ) +
  geom_hline(
    data = grp_stats,
    aes(yintercept = mean, color = age_d),
    linewidth = 2.5,
    linetype = "dashed"
  ) +
  geom_text(
    data = data.frame(
      x = mean(pltdta_dich$age[pltdta_dich$age_d == "age ≤60"]),
      y = grp_stats$mean[1] + 0.7,
      label = paste0("age ≤60 mean ≈ ", round(grp_stats$mean[1], 2)),
      age_d = "age ≤60"
    ),
    aes(x = x, y = y, label = label, color = age_d),
    hjust = 0,
    size = 4,
    inherit.aes = FALSE
  ) +
  geom_text(
    data = data.frame(
      x = mean(pltdta_dich$age[pltdta_dich$age_d == "age >60"]),
      y = grp_stats$mean[2] + 0.7,
      label = paste0("age >60 mean ≈ ", round(grp_stats$mean[2], 2)),
      age_d = "age >60"
    ),
    aes(x = x, y = y, label = label, color = age_d),
    hjust = 0,
    size = 4,
    inherit.aes = FALSE
  ) +
  geom_rect(
    data = data.frame(
      xmin = c(min(pltdta_dich$age[pltdta_dich$age_d == "age ≤60"]), 
               min(pltdta_dich$age[pltdta_dich$age_d == "age >60"])),
      xmax = c(max(pltdta_dich$age[pltdta_dich$age_d == "age ≤60"]), 
               max(pltdta_dich$age[pltdta_dich$age_d == "age >60"])),
      ymin = grp_stats$mean - grp_stats$sd,
      ymax = grp_stats$mean + grp_stats$sd,
      group_fill = c(col1, col2)
    ),
    aes(xmin = xmin, xmax = xmax, ymin = ymin, ymax = ymax, fill = group_fill),
    alpha = 0.3,
    inherit.aes = FALSE
  ) +
  scale_fill_manual(values = c("age ≤60" = col1, "age >60" = col2), guide = "none") +
  annotate(
    "text",
    x = mean(pltdta_dich$age[pltdta_dich$age_d == "age ≤60"]) - 1.5,
    y = grp_stats$mean[1] + grp_stats$sd[1],
    label = "±1 SD (≤60)",
    hjust = 1, color = "gray30", size = 3.5
  ) +
  annotate(
    "segment",
    x = mean(pltdta_dich$age[pltdta_dich$age_d == "age >60"]) - 1, 
    xend = mean(pltdta_dich$age[pltdta_dich$age_d == "age >60"]) - 0.5,
    y = grp_stats$mean[2] + grp_stats$sd[2],
    yend = grp_stats$mean[2] + grp_stats$sd[2],
    color = col2, linewidth = 3, alpha = 0.6
  ) +
  annotate(
    "text",
    x = mean(pltdta_dich$age[pltdta_dich$age_d == "age >60"]) - 1.5,
    y = grp_stats$mean[2] + grp_stats$sd[2],
    label = "±1 SD (>60)",
    hjust = 1, color = "gray30", size = 3.5
  ) +
  labs(x = "Age", y = "Score", title = "Dichotomized age model", color="") + 
  coord_cartesian(ylim = c(10, 20)) +
  guides(color = guide_legend(override.aes = list(line = 3)))
```

::: {.smaller}
Predictor $X_i$ on **x-axis** (age), observations on y-axis

* line represents group specific average
* greyed band +/- 1 group specific standard deviation
:::

:::

::: {.column width="5%"}

:::

::: {.column width="45%"}

Same model, but

- residual variance increases
- explained variance decreases
- residuals are clustered, not independent

violation of the distributional assumption:

$$
\varepsilon_i \overset{\text{i.i.d.}}{\sim} \mathcal{N}(0, \sigma^2).
$$


:::

```{r}
#| label: lm_dich
#| echo: true
lm_dich <- lm(score ~ age_d, data = dta_dich)
```

```{r}
#| label: tab-tidy_lm_dich
#| tab-cap: "Result dichomotomised predictor"
tidy_gt(lm_dich)
glance_gt(lm_dich)
```
::::

## Multiple Regression; combining information {data-toc-title="Multiple"}


A regression model with an intercept, a group difference, and a continuous predictor

- predictor specific means
- that can differ depending on other predictors
- with standard errors for each slope

The regression model starts being interesting:

$$
y_i = \beta_0  + \beta_1 x_{1i} + \beta_2 x_{2i} + \beta_3 x_{1i} x_{2i} + \varepsilon_i,
\qquad \text{for } i = 1,\dots,n,
$$

with the usual distributional assumption:

$$
\varepsilon_i \overset{\text{i.i.d.}}{\sim} \mathcal{N}(0, \sigma^2).
$$

&nbsp;

### Interaction

Interactions would allow effects to be condition dependent.  

- Age effect differs per group.
- Group difference are age dependent.

:::: {.columns}

::: {.column width="50%" .light-grey-col}

```{r}
#| label: fig-combined
#| fig-cap: "Interacting Predictors"
#| fig-height: 5.5
#| fig-width: 7
# use dta_int

lm_int <- lm(score ~ age * group, data = dta_int)

int_stats_overall <- dta_int |>
  summarise(
    mean_score = mean(score),
    sd_score   = sd(score),
    n          = n(),
    se_mean    = sd_score / sqrt(n()),
    .groups = "drop"
  )

pltdta_int <- dta_int |>
  mutate(
    fitted = predict(lm_int),
    resid  = score - fitted
  )

col_g0 <- col1
col_g1 <- col2

ggplot(pltdta_int, aes(x = age, y = score, color = group)) +
  geom_point(size = 2, alpha = 0.9) +
  geom_smooth(
    method = "lm", se = TRUE, 
    linewidth = 1.5, alpha = 0.2, fullrange = TRUE
  ) +
  geom_hline(
    data = int_stats_overall,
    aes(yintercept = mean_score),
    color = "gray60",
    linewidth = 2,
    linetype = "dashed",
    alpha = 0.8
  ) +
  geom_rect(
    data = data.frame(
      xmin = min(pltdta_int$age) - 1,
      xmax = max(pltdta_int$age) + 1,
      ymin = int_stats_overall$mean_score - int_stats_overall$sd_score,
      ymax = int_stats_overall$mean_score + int_stats_overall$sd_score
    ),
    aes(xmin = xmin, xmax = xmax, ymin = ymin, ymax = ymax),
    fill = "gray70",
    alpha = 0.15,
    inherit.aes = FALSE
  ) +
  annotate(
    "text",
    x = mean(pltdta_int$age),
    y = max(pltdta_int$score) - 0.4,
    label = paste0(
      "control: score ≈ ",
      round(coef(lm_int)[["(Intercept)"]], 2), " + ",
      round(coef(lm_int)[["age"]], 3), " × age"
    ),
    color = col_g0,
    size = 4
  ) +
  annotate(
    "text",
    x = mean(pltdta_int$age),
    y = max(pltdta_int$score) - 1.0,
    label = paste0(
      "treatment: score ≈ (",
      round(coef(lm_int)[["(Intercept)"]], 3), " ",
      round(coef(lm_int)[["grouptreatment"]], 3),") + (",
      round(coef(lm_int)[["age"]], 3), " ",
      round(coef(lm_int)[["age:grouptreatment"]], 3), ") × age"
    ),
    color = col_g1,
    size = 4
  ) +
  labs(
    x = "Age (continuous predictor)",
    y = "Score (outcome)",
    color = "Group",
    title = "Continuous predictor with group interaction"
  ) + 
  coord_cartesian(ylim = c(10, 20)) +
  scale_color_manual(values = c("control" = col_g0, "treatment" = col_g1)) + 
  theme(legend.position = "none")
```

::: {.smaller}
Predictor $x_{2i}$ on **x-axis** (age), observations on y-axis

* line represents group specific **linear trend**
* greyed band +/- 1 group specific standard deviation
:::

:::

::: {.column width="5%"}

:::

::: {.column width="45%"}

```{r}
#| label: lm_interact
#| echo: true
lm_int <- lm(score ~ age * group, data = dta_int)
```

```{r}
#| label: tab-tidy_lm_interact_estimates
#| tab-cap: "Result interaction effect"
tidy_gt(lm_int)
```

In this case, no evidence for an interaction.

:::

```{r}
#| label: tab-tidy_lm_interact_model
#| tab-cap: "Result interaction effect"
glance_gt(lm_int)
```

&nbsp;

::::

### Main effects

Simplifying the model and its interpretation, the interaction is removed.

When age is considered jointly with group.


:::: {.columns}

::: {.column width="50%" .light-grey-col}

```{r}
#| label: fig-combined-main
#| fig-cap: "Non-interacting Predictors"
#| fig-height: 5.5
#| fig-width: 7
# use dta_int

lm_int_main <- lm(score ~ age + group, data = dta_int)

int_stats_overall <- dta_int |>
  summarise(
    mean_score = mean(score),
    sd_score   = sd(score),
    n          = n(),
    se_mean    = sd_score / sqrt(n()),
    .groups = "drop"
  )

pltdta_int_main <- dta_int |>
  mutate(
    fitted = predict(lm_int_main),
    resid  = score - fitted
  )

col_g0 <- col1
col_g1 <- col2

ggplot(pltdta_int_main, aes(x = age, y = score, color = group)) +
  geom_point(size = 2, alpha = 0.9) +
  geom_smooth(
    method = "lm", se = TRUE, 
    linewidth = 1.5, alpha = 0.2, fullrange = TRUE
  ) +
  geom_hline(
    data = int_stats_overall,
    aes(yintercept = mean_score),
    color = "gray60",
    linewidth = 2,
    linetype = "dashed",
    alpha = 0.8
  ) +
  geom_rect(
    data = data.frame(
      xmin = min(pltdta_int_main$age) - 1,
      xmax = max(pltdta_int_main$age) + 1,
      ymin = int_stats_overall$mean_score - int_stats_overall$sd_score,
      ymax = int_stats_overall$mean_score + int_stats_overall$sd_score
    ),
    aes(xmin = xmin, xmax = xmax, ymin = ymin, ymax = ymax),
    fill = "gray70",
    alpha = 0.15,
    inherit.aes = FALSE
  ) +
  annotate(
    "text",
    x = mean(pltdta_int_main$age),
    y = max(pltdta_int_main$score) - 0.4,
    label = paste0(
      "group 0: score ≈ ",
      round(coef(lm_int_main)[["(Intercept)"]], 2), " + ",
      round(coef(lm_int_main)[["age"]], 3), " × age"
    ),
    color = col_g0,
    size = 4
  ) +
  annotate(
    "text",
    x = mean(pltdta_int_main$age),
    y = max(pltdta_int_main$score) - 1.0,
    label = paste0(
      "group 1: score ≈ (",
      round(coef(lm_int_main)[["(Intercept)"]], 3), " ",
      round(coef(lm_int_main)[["grouptreatment"]], 3),") + ",
      round(coef(lm_int_main)[["age"]], 3), " x age"
    ),
    color = col_g1,
    size = 4
  ) +
  labs(
    x = "Age (continuous predictor)",
    y = "Score (outcome)",
    color = "Group",
    title = "Continuous predictor without group interaction"
  ) + 
  coord_cartesian(ylim = c(10, 20)) +
  scale_color_manual(values = c("control" = col_g0, "treatment" = col_g1)) + 
  theme(legend.position = "none")
```

::: {.smaller}
Predictor $x_{2i}$ on **x-axis** (age), observations on y-axis

* line represents group specific **linear trend**
* greyed band +/- 1 group specific standard deviation
:::

:::

::: {.column width="5%"}

:::

::: {.column width="45%"}

```{r}
#| label: lm_main
#| echo: true
lm_int_main <- lm(score ~ age + group, data = dta_int)
```

```{r}
#| label: tab-tidy_lm_main_estimates
#| tab-cap: "Result 2 main effects"
tidy_gt(lm_int_main)
```

There is a very clear group effect.
In this case, no evidence for an interaction.

:::

```{r}
#| label: tab-tidy_lm_main_model
#| tab-cap: "Result 2 main effects"
glance_gt(lm_int_main)
```

&nbsp;

::::

### Ignoring a relevant predictor

With only group in the model, ignoring age.


:::: {.columns}

::: {.column width="50%" .light-grey-col}

```{r}
#| label: fig-ignored
#| fig-cap: "Ignoring a predictor"
#| fig-height: 5.5
#| fig-width: 7
# use dta_int

lm_int_ign <- lm(score ~ group, data = dta_int)

int_stats_overall <- dta_int |>
  summarise(
    mean_score = mean(score),
    sd_score   = sd(score),
    n          = n(),
    se_mean    = sd_score / sqrt(n()),
    .groups = "drop"
  )

pltdta_int_ign <- dta_int |>
  mutate(
    fitted = predict(lm_int_ign),
    resid  = score - fitted
  )

col_g0 <- col1
col_g1 <- col2

ggplot(pltdta_int_ign, aes(x = group, y = score, color = group)) +
  geom_jitter(size = 2, alpha = 0.9, width=.2) +
  geom_smooth(
    method = "lm", se = TRUE, 
    linewidth = 1.5, alpha = 0.2, fullrange = TRUE
  ) +
  geom_hline(
    data = int_stats_overall,
    aes(yintercept = mean_score),
    color = "gray60",
    linewidth = 2,
    linetype = "dashed",
    alpha = 0.8
  ) +
  labs(
    x = "Groups (categorical predictor)",
    y = "Score (outcome)",
    color = "Group",
    title = "Ignoring continuous predictor"
  ) + 
  coord_cartesian(ylim = c(10, 20)) +
  scale_color_manual(values = c("control" = col_g0, "treatment" = col_g1)) + 
  theme(legend.position = "none")
```

::: {.smaller}
Predictor $x_{1i}$ on **x-axis** (age), observations on y-axis
:::

:::

::: {.column width="5%"}

:::

::: {.column width="45%"}

```{r}
#| label: lm_group_only
#| echo: true
lm_int_ign <- lm(score ~ group, data = dta_int)
```

```{r}
#| label: tab-tidy_lm_ignored
#| tab-cap: "Result regression on groups only"
tidy_gt(lm_int_ign)
```

There is no group effect !  
Remember: conditional on age there is one.

Do not make decisions based on univariable analysis unless you have to.


:::

```{r}
#| label: tab-tidy_lm_ignored_model
#| tab-cap: "Result ignoring effect"
glance_gt(lm_int_ign)
```

&nbsp;

::::

## Assumptions {data-toc-title="Assumptions"}

Models are based on assumptions,  
violations can compromise inference.

For linear regression the assumptions are:

1. **Linearity**: Predictor-outcome relationship is linear
2. **Independence**: Residuals are i.i.d. (no clustering/autocorrelation)
3. **Homoscedasticity**: Constant residual variance
4. **Normality**: Residuals normally distributed (for inference)

Evaluation of model assumption is based on (standardized) residuals.

To evaluate linearity, plot the outcome in relation to

- the fitted value for the whole model
- individual predictors for parts of the model

To evaluate independence, plot the residuals in relation to

- the fitted value for the whole model
- individual predictors for parts of the model

and watch out for patterns

To evaluate homoscedasticity, plot the standardized residuals in relation to

- individual predictors for parts of the model

and watch out for predictor specific spread

```{r}
#| label: assumptions_prep
assumpt <- dta_int |> 
  mutate(
    fitted = fitted(lm_int),
    resid = residuals(lm_int),
    resid_std = rstandard(lm_int),
    theoretical = rnorm(length(resid))
  ) 
```

```{r}
#| label: fig-residual_analysis
#| fig-cap: "Residual Structure"
#| fig-height: 5.5
#| fig-width: 12
plt_resid_fitted <- ggplot(data=assumpt, aes(x = fitted, y = resid)) +
  geom_point() +
  geom_hline(yintercept = 0, linetype = "dashed") +
  # theme_classic() +
  labs(x = "Fitted values", y = "Residuals", title = "Residuals vs fitted")

plt_resid_age <- ggplot(data=assumpt, aes(x = age, y = resid)) +
  geom_point() +
  geom_hline(yintercept = 0, linetype = "dashed") +
  # theme_classic() +
  labs(x = "Age", y = "Residuals", title = "Residuals vs age")

plt_resid_group <- ggplot(data=assumpt, aes(x = group, y = resid)) +
  geom_boxplot() +
  # theme_classic() +
  labs(x = "Group", y = "Residuals", title = "Residuals by group")

wrap_plots(plt_resid_fitted,plt_resid_age,plt_resid_group, nrow=1)
```

To evaluate normality of the residuals, use a qqplot.  
Note, normality of (standardized) residuals follows from the assumption that disturbances are random and additive.

```{r}
#| label: fig-residual_normality
#| fig-cap: "Residual Normality"
#| fig-height: 5.5
#| fig-width: 10
plt_resid_normal <- ggplot(data=assumpt, aes(sample = resid_std)) +
  stat_qq(distribution = qnorm, color = col1) +
  stat_qq_line(distribution = qnorm, color = col2) +
  labs(title = "Q-Q Plot: Residuals vs Theoretical Normal") 

plt_resid_hist <- ggplot(data=assumpt) + 
  geom_histogram(aes(x = resid, y = after_stat(density)), 
                bins = 15, alpha = 0.7, fill = col2) +
  geom_density(aes(x = resid), color = col1, linewidth = 1) +
  labs(title = "Residuals Histogram + Density")

wrap_plots(plt_resid_normal,plt_resid_hist, nrow=1)
```

For the identically and individually distributed assumption:

$$\varepsilon_i \overset{\text{i.i.d.}}{\sim} \mathcal{N}(0, \sigma^2). $$

The lack of any patterns supports the i.i.d. assumption (does not prove it)
Remaining patterns in the residuals can possibly be addressed

- using additional variables
- using transformations

Build models to capture the systematicity in the data,  
so the remaining part is random, and ideally relatively small.

For completeness,  
while tests are just tests:

```{r}
#| label: tab-assumption_summary
#| tab-cap: "Diagnostics Tests"
tests <- data.frame(
  Test = c("Durbin-Watson (Independence)", "Breusch-Pagan (Homoscedasticity)", "Shapiro-Wilk (Normality)"),
  Result = c(
    round(car::durbinWatsonTest(lm_int)$dw, 2),
    round(car::ncvTest(lm_int)$p, 3),
    round(shapiro.test(residuals(lm_int))$p.value, 3)
  ),
  Interpretation = c(
    "≈2 = no autocorrelation ✓",
    "p > 0.05 = constant variance ✓", 
    "p > 0.05 = normality ✓"
  )
)
knitr::kable(tests)
```

```{r}
save.image('zenzi.RData')
```