# Longitudinal Data

When clustering is because of repeated measurements over time:

- same mixed model dependencies arise within units 
- but **time is ordered**.
- temporal structure: measurements closer in time could be more similar

We now re-interpret the two levels:

- **Level 1**: observations (at different time points), within patient
- **Level 2**: patient

The random effects shape individual **trajectories**.

```{r}
dta_timed <- dta_grouped |> group_by(group) |> arrange(age) |> mutate(time=age,timec=factor(1:n())) |> select(-age) |> ungroup()
set.seed(123)
.devs <- abs(rnorm(9,0,.5))
.res <- rnorm(64,0,2)
dta_timed <- dta_timed |> mutate(score_x = score + .devs[group]*time + .res)
# maybe x2 not required
dta_timed_x2 <- dta_grouped_x2 |> group_by(group) |> mutate(time=age,timec=1:n()) |> select(-age)
set.seed(321)
dta_timed_x2 <- dta_timed_x2 |> mutate(score=score+rnorm(n(),0,2))
dta_timed_x2 <- dta_timed_x2 |>
  mutate(time_c = scale(time, center = TRUE, scale = FALSE))
```

## Categorical and Continuous Time Scale  {data-toc-title="Time Scales"}

Time within mixed models

- does not require 
  - equidistance
  - fixed timepoints
- can be categorical (ordinal) or continuous (numeric)

**Choose continuous** when:

- time is meaningful numerically; time elapsed is a driver
- trajectories behave smooth over time

**Choose categorical** when:

- time is not fully meaningful numerically
  - order can be a driver
  - trajectories change in nature (data generating process) 
  - e.g., pre-post, 3m, 6m, 12m

```{r}
#| label: prep_fig_continuous_trajectories
#| echo: false
plt_cnt_time <- ggplot(dta_timed, aes(x=time, y=score, group = group, color = group)) +
  geom_smooth(method='lm', se=FALSE) +
  geom_line(alpha=.5) +
  geom_point(size = 2.5) +
  scale_color_manual(
    values = ifelse(levels(dta_timed$group) == "5", "red", "black")
  ) +
  guides(color = "none")
```

```{r}
#| label: prep_fig_categorical_trajectories
#| echo: false
plt_cat_time <- ggplot(dta_timed, aes(x=timec, y=score, group = group, color = group)) +
  geom_smooth(method='lm', se=FALSE) +
  geom_line(alpha=.5) +
  geom_point(size = 2.5) +
  scale_color_manual(
    values = ifelse(levels(dta_timed$group) == "5", "red", "black")
  ) +
  guides(color = "none")
```

```{r}
#| label: fig-categorical_continuous
#| fig-cap: "Time Scales: Continuous vs Categorical"
#| fig-height: 5.5
#| fig-width: 12
#| echo: false
wrap_plots(plt_cnt_time, plt_cat_time, nrow=1)
```

## The Longitudinal Random Intercept Model  {data-toc-title="Random Intercept"}

The simplest longitudinal model

- individuals start at different baselines
- change at the same average rate
- just a random intercept model with time as predictor
- correlation within clusters, independent of time difference

::: {.equation-box}
$$
y_{it} = \underbrace{(\gamma_{00} + u_{0i})}_{\text{person-specific intercept}} + \underline{\gamma_{10} \cdot \text{time}_{it}} + \varepsilon_{it}
$$

with:
$$
u_{0i} \sim \mathcal{N}(0,\ \tau_{00}), \qquad \varepsilon_{it} \sim \mathcal{N}(0,\ \sigma^2)
$$
:::


The model fit now uses time as fixed variable to capture the average change,  
while accomodating a constant covariance between residuals remaining.

```{r}
#| label: long_ri_9_prep
dta_timed_x <- dta_timed |> mutate(score=score_x)
```

```{r}
#| label: long_ri_9
#| echo: true
#| fig-height: 7
#| fig-width: 10
lmert_int_9 <- lmer(score ~ time + (1 | group), data = dta_timed_x)
```

```{r}
#| label: prep_fig_random_intercept_time
#| fig-height: 5.5
#| fig-width: 5
# lmert_int_9 <- lmer(score ~ time + (1 + time | group), data = dta_timed)
pltdta_ri_time <- augment(lmert_int_9, data = dta_timed_x) # |>  mutate(group = factor(group))

ranefs_df <- ranef(lmert_int_9)$group |>
  as_tibble(rownames = "group") |>
  mutate(group = factor(group))

# Create group-specific fitted lines (same slope, different intercepts)
pred_data_groups <- ranefs_df |>
  select(group, `(Intercept)`) |>
  crossing(
    time = seq(
      min(pltdta_ri_time$time, na.rm = TRUE),
      max(pltdta_ri_time$time, na.rm = TRUE),
      length.out = 200
    )
  ) |>
  mutate(
    fitted = fixef(lmert_int_9)[["(Intercept)"]] +
             `(Intercept)` +
             fixef(lmert_int_9)[["time"]] * time
  )

# Population-level fixed effect line
pred_data_pop <- tibble(
  time = seq(
    min(pltdta_ri_time$time, na.rm = TRUE),
    max(pltdta_ri_time$time, na.rm = TRUE),
    length.out = 200
  )
) |>
  mutate(
    fitted = fixef(lmert_int_9)[["(Intercept)"]] +
             fixef(lmert_int_9)[["time"]] * time
  )

time_stats <- pltdta_ri_time |>
  summarise(
    mean_score = mean(score, na.rm = TRUE),
    se_mean = sd(score, na.rm = TRUE) / sqrt(sum(!is.na(score)))
  )

n_groups <- nlevels(pltdta_ri_time$group)
#group_grays_9 <- gray.colors(n_groups, start = 0.25, end = 0.85)
#col2 <- "#5B8FA8"
```

```{r}
#| label: fig-random_intercept_time
#| fig-cap: "Random intercept over time"
#| fig-height: 5.5
#| fig-width: 5
ggplot(pltdta_ri_time, aes(x = time, y = score)) +
  geom_point(
    aes(color = group),
    size = 2.5,
    alpha = 0.85
  ) +
  scale_color_manual(values = group_grays_9, guide = "none") +
  geom_segment(
    aes(x = time, xend = time, y = score, yend = .fitted),
    color = "gray60",
    alpha = 0.30,
    linewidth = 0.8
  ) +
  geom_line(
    data = pred_data_pop,
    aes(x = time, y = fitted),
    inherit.aes = FALSE,
    color = "black",
    linewidth = 1.2
  ) +
 geom_line(
    data = pred_data_groups,
    aes(x = time, y = fitted, group = group, color = group),
    linewidth = 0.9,
    alpha = 0.7
  ) +
  annotate(
    "text",
    x = mean(pltdta_ri_time$time, na.rm = TRUE),
    y = max(pltdta_ri_time$score, na.rm = TRUE) + 0.5,
    label = paste0(
      "score ≈ ",
      round(fixef(lmert_int_9)[["(Intercept)"]], 2),
      " + ",
      round(fixef(lmert_int_9)[["time"]], 3),
      " × time + random intercept(group)"
    ),
    size = 4.3
  ) +
  labs(
    x = "Time (continuous)",
    y = "Score",
    title = "Random Intercept Model",
    subtitle = "Parallel group lines: same slope, different intercepts"
  ) +
  theme_lmm() +
  theme(
    plot.title = element_text(face = "bold"),
    plot.subtitle = element_text(size = 11, color = "gray40"),
    panel.grid.minor = element_blank()
  ) +
  coord_cartesian(
    xlim = c(
      min(pltdta_ri_time$time, na.rm = TRUE),
      max(pltdta_ri_time$time, na.rm = TRUE) + 0.12 * diff(range(pltdta_ri_time$time, na.rm = TRUE))
    )
  )
```


### Categorical Time

With 4 time points (**T0, T3, T6, T9**), we can compare scores in two ways:

1. **Consecutive timepoints**: T3 vs T0, T6 vs T3, T9 vs T6 (differences between adjacent times)
2. **All vs reference**: T3 vs T0, T6 vs T0, T9 vs T0 (all times compared to baseline)

```{r}
#| label: setup_categorical_time
dta_timed_c <- dta_timed |> filter(timec %in% c(1,2,3,4))

dta_timed_c$timec <- factor(dta_timed_c$timec, labels = c("T0", "T3", "T6", "T9"), levels = c(1, 2, 3, 4))
```

Expected marginal means used to extract contrasts from a fitted model:

```{r}
#| label: emmeans_setup
model_em <- lmer(score ~ timec + (1 | group), data = dta_timed_c)
```

All vs Reference (T0)

```{r}
#| label: emmeans_all_vs_ref
em_all_vs_ref <- emmeans(model_em, specs = "timec")
tidy(em_all_vs_ref)
tidy(contrast(em_all_vs_ref, method = "trt.vs.ctrl", ref = 1))
```

Consecutive Comparisons

```{r}
#| label: emmeans_consecutive
tidy(contrast(em_all_vs_ref, method = "consec"))
```

All Pairwise Comparisons

```{r}
#| label: emmeans_pairwise
contrast(em_all_vs_ref, method = "pairwise")
```

Custom Comparisons can address specific research questions,  
E.g., T9 vs avg(T3, T6) uses contrast parameters 0, -0.5, -0.5, 1

```{r}
contrast(em_all_vs_ref, method = list('T9 - (T6+T3)/2'=c(0, -0.5, -0.5, 1)))
```

## The Longitudinal Random Intercept - Random Slope Model  {data-toc-title="Random Slope"}

Heterogeneity can exist over intercept, but also over slopes

- individuals differ in slope
- with an average slope, and an individual specific difference from that average
- time as fixed and random effect

For longitudinal data, individuals can differ in their average but also in their change over time.

::: {.equation-box}
$$
y_{it} = (\gamma_{00} + u_{0i}) + (\gamma_{10} + u_{1i}) * \text{time}_{it} + \varepsilon_{it}
$$

with:
$$
\begin{pmatrix} u_{0i} \\ u_{1i} \end{pmatrix} \sim \mathcal{N}\!\left(\mathbf{0},\ \begin{pmatrix} \tau_{00} & \tau_{01} \\ \tau_{01} & \tau_{11} \end{pmatrix}\right), \qquad \varepsilon_{it} \sim \mathcal{N}(0,\ \sigma^2)
$$
:::

The covariance $\tau_{01}$ is often negative, starting higher results in stronger decrease, or starting lower in stronger increase.

Again: To avoid convergence issues you should center the random slope to reduce colinearity with the intercept.

```{r}
#| label: long_rs_9
#| echo: true
lmert_slp_9 <- lmer(score ~ time + (1 + time | group), data = dta_timed_x)
```

```{r}
#| label: long_rs_9_sum
tidy_gt(lmert_slp_9)
```

```{r}
#| label: prep_fig_random_slope_time
#| fig-height: 5.5
#| fig-width: 5
# lmert_int_9 <- lmer(score ~ time + (1 + time | group), data = dta_timed)
pltdta_rs_time <- augment(lmert_slp_9, data = dta_timed_x) # |>  mutate(group = factor(group))

ranefs_df <- ranef(lmert_slp_9)$group |>
  as_tibble(rownames = "group") |>
  mutate(group = factor(group))

# Create group-specific fitted lines (same slope, different intercepts)
pred_data_groups <- ranefs_df |>
  select(group, `(Intercept)`, rtime = `time`) |>
  crossing(
    time = seq(
      min(pltdta_rs_time$time, na.rm = TRUE),
      max(pltdta_rs_time$time, na.rm = TRUE),
      length.out = 200
    )
  ) |>
  mutate(
    fitted = fixef(lmert_slp_9)[["(Intercept)"]] +
             `(Intercept)` +
             (fixef(lmert_slp_9)[["time"]] + rtime)* time
  )

# Population-level fixed effect line
pred_data_pop <- tibble(
  time = seq(
    min(pltdta_rs_time$time, na.rm = TRUE),
    max(pltdta_rs_time$time, na.rm = TRUE),
    length.out = 200
  )
) |>
  mutate(
    fitted = fixef(lmert_slp_9)[["(Intercept)"]] +
             fixef(lmert_slp_9)[["time"]] * time
  )

time_stats <- pltdta_rs_time |>
  summarise(
    mean_score = mean(score, na.rm = TRUE),
    se_mean = sd(score, na.rm = TRUE) / sqrt(sum(!is.na(score)))
  )

n_groups <- nlevels(pltdta_rs_time$group)
#group_grays_9 <- gray.colors(n_groups, start = 0.25, end = 0.85)
#col2 <- "#5B8FA8"
```

```{r}
#| label: fig-random_slope_time
#| fig-cap: "Random slope model over time"
#| fig-height: 5.5
#| fig-width: 5
ggplot(pltdta_rs_time, aes(x = time, y = score)) +
  geom_point(
    aes(color = group),
    size = 2.5,
    alpha = 0.85
  ) +
  scale_color_manual(values = group_grays_9, guide = "none") +
  geom_segment(
    aes(x = time, xend = time, y = score, yend = .fitted),
    color = "gray60",
    alpha = 0.30,
    linewidth = 0.8
  ) +
  geom_line(
    data = pred_data_pop,
    aes(x = time, y = fitted),
    inherit.aes = FALSE,
    color = "black",
    linewidth = 1.2
  ) +
 geom_line(
    data = pred_data_groups,
    aes(x = time, y = fitted, group = group, color = group),
    linewidth = 0.9,
    alpha = 0.7
  ) +
  annotate(
    "text",
    x = mean(pltdta_rs_time$time, na.rm = TRUE),
    y = max(pltdta_rs_time$score, na.rm = TRUE) + 0.5,
    label = paste0(
      "score ≈ ",
      round(fixef(lmert_slp_9)[["(Intercept)"]], 2),
      " + ",
      round(fixef(lmert_slp_9)[["time"]], 3),
      " × time + random effects(group)"
    ),
    size = 4.3
  ) +
  labs(
    x = "Time (continuous)",
    y = "Score",
    title = "Random Intercept, Random Slope Model",
    subtitle = "Non-parallel group lines: different slopes, different intercepts"
  ) +
  theme_lmm() +
  theme(
    plot.title = element_text(face = "bold"),
    plot.subtitle = element_text(size = 11, color = "gray40"),
    panel.grid.minor = element_blank()
  ) +
  coord_cartesian(
    xlim = c(
      min(pltdta_rs_time$time, na.rm = TRUE),
      max(pltdta_rs_time$time, na.rm = TRUE) + 0.12 * diff(range(pltdta_rs_time$time, na.rm = TRUE))
    )
  )
```

# Covariance Structures

Whithin cluster correlational structures:

- correlation same for all pairs
- correlation different for all pairs
- correlation structured (in between the above)

Choose **covariance structure** for the outcome vector $\mathbf{y}_i$ of patient $i$.

Note: covariance structures 

- implied by the variance-covariance estimated
- specified as a repeated measures structure (residual covariance).

For $T = 4$ time points, the covariance matrix $\boldsymbol{\Sigma}_i$ is 5×5. 

## Most Common Structures {data-toc-title="Examples"}

### Compound Symmetry (CS)

All variances equal; all covariances equal regardless of time lag.

$$
\boldsymbol{\Sigma}_{\text{CS}} =
\begin{pmatrix}
\sigma^2 & \rho\sigma^2 & \rho\sigma^2 & \rho\sigma^2 \\
\rho\sigma^2 & \sigma^2 & \rho\sigma^2 & \rho\sigma^2 \\
\rho\sigma^2 & \rho\sigma^2 & \sigma^2 & \rho\sigma^2 \\
\rho\sigma^2 & \rho\sigma^2 & \rho\sigma^2 & \sigma^2
\end{pmatrix}
$$

**This is exactly what a random intercept model implies.** Two parameters: $\sigma^2$ and $\rho$.

Note: this is often used for clustered data, not necessarily repeated measurements.

### Unstructured (UN)

No constraints — every variance and every covariance is estimated freely. For $T$ time points: $T(T+1)/2$ parameters. Most flexible, but most expensive in terms of parameters.

$$
\boldsymbol{\Sigma}_{\text{Unstructured}} =
\begin{pmatrix}
\sigma_1^2 & \sigma_{12} & \sigma_{13} & \sigma_{14} \\
\sigma_{12} & \sigma_2^2 & \sigma_{23} & \sigma_{24} \\
\sigma_{13} & \sigma_{23} & \sigma_3^2 & \sigma_{34} \\
\sigma_{14} & \sigma_{24} & \sigma_{34} & \sigma_4^2
\end{pmatrix}
$$

### Heterogeneous Compound Symmetry (CSH)

Like CS, but allows variances to differ across time points. Useful when variability grows over follow-up.

$$
\boldsymbol{\Sigma}_{\text{CSH}} =
\begin{pmatrix}
\sigma_1^2 & \rho\,\sigma_1\sigma_2 & \rho\,\sigma_1\sigma_3 & \rho\,\sigma_1\sigma_4 \\
\rho\,\sigma_1\sigma_2 & \sigma_2^2 & \rho\,\sigma_2\sigma_3 & \rho\,\sigma_2\sigma_4 \\
\rho\,\sigma_1\sigma_3 & \rho\,\sigma_2\sigma_3 & \sigma_3^2 & \rho\,\sigma_3\sigma_4 \\
\rho\,\sigma_1\sigma_4 & \rho\,\sigma_2\sigma_4 & \rho\,\sigma_3\sigma_4 & \sigma_4^2
\end{pmatrix}
$$

### Autoregressive Order 1 — AR(1)

Correlation decays exponentially with lag: $\text{Corr}(y_t, y_{t+k}) = \rho^k$.

$$
\boldsymbol{\Sigma}_{\text{AR(1)}} =
\sigma^2
\begin{pmatrix}
1 & \rho & \rho^2 & \rho^3 \\
\rho & 1 & \rho & \rho^2 \\
\rho^2 & \rho & 1 & \rho \\
\rho^3 & \rho^2 & \rho & 1
\end{pmatrix}
$$

Two parameters: $\sigma^2$ and $\rho$. Natural for equally-spaced measurements with decaying memory.

### Heterogeneous AR(1)

Allow the variances to change over time.

$$
\boldsymbol{\Sigma}_{\text{ARH(1)}} =
\begin{pmatrix}
\sigma_1^2 & \rho\,\sigma_1\sigma_2 & \rho^2\,\sigma_1\sigma_3 & \rho^3\,\sigma_1\sigma_4 \\
\rho\,\sigma_1\sigma_2 & \sigma_2^2 & \rho\,\sigma_2\sigma_3 & \rho^2\,\sigma_2\sigma_4 \\
\rho^2\,\sigma_1\sigma_3 & \rho\,\sigma_2\sigma_3 & \sigma_3^2 & \rho\,\sigma_3\sigma_4 \\
\rho^3\,\sigma_1\sigma_4 & \rho^2\,\sigma_2\sigma_4 & \rho\,\sigma_3\sigma_4 & \sigma_4^2
\end{pmatrix}
$$

### Toeplitz

Correlations depend only on the lag, but without the exponential constraint of AR(1). Each lag distance has its own parameter. More flexible than AR(1), fewer parameters than unstructured.

$$
\boldsymbol{\Sigma}_{\text{Toeplitz}} =
\begin{pmatrix}
\tau_0 & \tau_1 & \tau_2 & \tau_3 \\
\tau_1 & \tau_0 & \tau_1 & \tau_2 \\
\tau_2 & \tau_1 & \tau_0 & \tau_1 \\
\tau_3 & \tau_2 & \tau_1 & \tau_0
\end{pmatrix}
$$

### 6. ...

Various others exist, choose 

- most efficient
- captures the assumed process
  - continuous or not, equidistant or not ?
- is in line with the data
  - choose based on AIC or BIC (fitted with ML)


```{r}
save.image('zenzi.RData')
```



