## Adding a Random Slope {data-toc-title="Random Slope"}

### Heterogeneity of Effects

A cluster specific slope (effect) can be considered sampled from a common population, with a mean and variance. This is equivalent to cluster specific intercepts.

::: {.equation-box}
**Level 1:**
$$
y_{ij} = \beta_{0j} + \underline{\beta_{1j} x_{ij}} + \varepsilon_{ij}, \qquad \varepsilon_{ij} \sim \mathcal{N}(0,\ \sigma^2)
$$

**Level 2:**
$$
\beta_{0j} = \gamma_{00} + u_{0j}, \qquad \underline{\beta_{1j}} = \gamma_{10} + u_{1j}
$$

**Distributional assumption for the random effects:**
$$
\begin{pmatrix} u_{0j} \\ u_{1j} \end{pmatrix} \sim \mathcal{N}\!\left(\begin{pmatrix}0\\0\end{pmatrix},\ \boldsymbol{\Psi}\right), \qquad \boldsymbol{\Psi} = \begin{pmatrix} \underline{\tau_{00}} & \underline{\tau_{01}} \\ \tau_{01} & \underline{\tau_{11}} \end{pmatrix}
$$

**Combined:**
$$
y_{ij} = \underbrace{\gamma_{00} + \gamma_{10} x_{ij}}_{\text{fixed}} + \underbrace{u_{0j} + u_{1j} x_{ij} + \varepsilon_{ij}}_{\text{random}}
$$
:::

$\boldsymbol{\Psi}$ captures:

- $\tau_{00}$: how much cluster intercepts vary
- $\tau_{11}$: how much cluster slopes vary
- $\tau_{01}$: covariance between intercept and slope 
  (e.g., high-baseline clusters may also have steeper slopes)

```{r}
#| eval: false
dta2 <- dta_grouped |> group_by(group) |> mutate(av=mean(score)) |> ungroup() |> mutate(av=av-mean(score),agec=age-mean(age)) |> mutate(score=score+(av*.05*agec))
dta2x <- bind_rows(dta_grouped, dta2 |>select(-c(av,agec)) |> mutate(group=group+10))
lmer(score~age+(1+age|group),data=dta2x)
ggplot(data=dta2x,aes(y=score,x=age,group=group)) + geom_line() + geom_point()
```


```{r}
#| label: dta_grouped_x2
dta_grouped_x2 <- dta_grouped |> group_by(group) |> mutate(av=mean(score)) |> ungroup() |> mutate(av=av-mean(score),agec=age-mean(age)) |> mutate(score=score+(av*.05*agec))
dta_grouped_x2 <- bind_rows(dta_grouped, dta_grouped_x2 |>select(-c(av,agec)) |> mutate(group=factor(as.numeric(as.character(group))+10)))
# ggplot(data=dta_grouped_x2,aes(y=score,x=age,group=group)) + geom_line() + geom_point()
```

```{r}
#| label: fig-random-slope
#| fig-cap: "Random intercept + random slope model."
#| fig-height: 5
ggplot(dta_grouped_x2, aes(x=age, y=score, color = factor(group))) +
  geom_point(alpha = 0.25, size = 1.5) + geom_smooth(method='lm', se=FALSE) + geom_smooth(aes(group=1), method='lm', color='black', linetype=2) +
  scale_color_manual(values=gray.colors(18,start=0.2,end=0.8), guide="none") + labs(title = "Increased variability of slopes",
       x = "age", y = "score", color = "Group")
```

```{r}
lmer_slope_18 <- lmer(score~age+(1+age|group),data=dta_grouped_x2)
tidy_mixed_gt(lmer_slope_18)
glance_mixed_gt(lmer_slope_18)
```

Convergence problems: 

- few groups (< 10–20), covariance $\boldsymbol{\Psi}$ can be hard to estimate reliably
  - center the predictors to remove correlation random intercept - random slope
  - avoid estimation of covariance: `(1 + age || group)` which constrains $\tau_{01} = 0$.

```{r}
tidy_mixed_gt(lmer(score~age+(1+age||group),data=dta_grouped))
```

# Technicalities

Statistical inference is about

- statistical estimation &rarr; size
  - estimate: my best guess
  - standard error: uncertainty about my best guess
- statistical testing &rarr; decision
  - p-value: probability that by chance
  - change in AIC: improved model

## ML vs REML in Mixed Models  {data-toc-title="ML vs REML"}

Estimation typically by **Maximum Likelihood (ML)** or **Restricted Maximum Likelihood (REML)**.  

Optimize likelihood function

- ML estimates **fixed effects and variance components together**
- REML estimate **the variance components** on the information after removal of fixed effects

REML reduces small-sample bias in variance estimates (small number of clusters)

Use **REML** when

- estimation of **variance components / random effects**.
- fitting a *final* model

Use **ML** when

- comparison of **nested models with different fixed-effects structure** 
- focus on testing or selecting fixed effects

## Testing Variance Components  {data-toc-title="Testing Variances"}

For variance components, the null hypothesis typically assumes variances at the **boundary** of the parameter space (e.g. $\tau^2 = 0$).  

The asymptotic distribution of the **likelihood ratio test (LRT)** can be considered a **mixture of chi-squares**.

### Example: testing a random intercept (single variance)

Null hypothesis: no between-cluster variability / heterogeneity

Test $H_0: \tau_{00} = 0$ vs $H_1: \tau_{00} > 0$.

Fit both models by **ML** (not REML)  

```{r}
#| label: lrt_single_variance
# H1: random intercept model
mod1_ml <- lmer(score ~ age + (1 | group),
                data = dta_grouped,
                REML = FALSE)

# H0: no random intercept (ordinary regression with group-ignored random part)
mod0_ml <- gls(score ~ age, data = dta_grouped)

# Put on a comparable ML deviance scale
logLik_mod1 <- logLik(mod1_ml)
logLik_mod0 <- logLik(mod0_ml)

LR_stat <- 2 * (as.numeric(logLik_mod1) - as.numeric(logLik_mod0))
```

The Likelihood Ratio test statistics is computed

```{r}
#| eval: false
#| echo: true
LR_stat <- 2 * (as.numeric(logLik_mod1) - as.numeric(logLik_mod0))
```

Under $H_0$, the variance component is on the boundary ($\tau_{00} = 0$),  the LRT does **not** follow $\chi^2_1$, but rather a **mixture** of chi-square distributions:

$\frac{1}{2} \chi^2_0 + \frac{1}{2} \chi^2_1$

where $\chi^2_0$ is a point mass at 0.  

This leads to a simple p-value adjustment: take half !

Evaluate the Likelihood ratio test statistics:

$p = 0.5 \cdot P(\chi^2_1 \ge \text{LR})$

```{r}
#| eval: false
#| label: lrt_single_variance_pvalue
# mixture 0.5 * chi^2_0 + 0.5 * chi^2_1
p_mix <- 0.5 * (1 - pchisq(LR_stat, df = 1))
p_naive <- 1 - pchisq(LR_stat, df = 1)

c(LR = LR_stat,
  p_mixture = p_mix,
  p_naive_chisq1 = p_naive)
```

### Beyond a single variance

Testing more complex hypothesis,  
about multiple variances and/or covariances

- e.g., testing a **random slope variance**
- e.g., testing a **variance and a covariance simultaneously**

asymptotic LRT distribution remains **finite mixture of chi-squared distributions**,  
with **degrees of freedom** and **weights** in the mixture depend on 

  - number of constrained variance and covariance parameters
  - structure of the covariance matrix $H_0$ and $H_1$

More complex models require simulation.

Note: testing covariance is not a challenge

```{r}
save.image('zenzi.RData')
```