# Dependent data

Data is often internally dependent:

- cross-sectional clustering: groups of observations (classes, litters, clinics, ...)
- within-unit clustering: repeated measures (trajectories, pre-post, ...)
- distances (less relation when further)
- study effects measuring the same (meta-analysis)
- spatial random effects
- item response models, with respondent clustering
- ...

What dependency looks like:

```{r}
#| label: dta_group9_by_residuals_and_random
# lm_cont <- lm(score ~ age, data = dta_cont) # use dta_cont, dta_grouped
pltdta_cont <- dta_cont
pltdta_cont$resid <- residuals(lm_cont)
dta_sorted <- pltdta_cont[order(-pltdta_cont$resid), ]
dta_sorted$group <- factor(rep(1:9, times = c(4, 4, 8, 8, 16, 8, 8, 4, 4)))
dta_grouped <- dta_sorted[order(dta_sorted$id), ]
set.seed(123)
dta_grouped$group0 <- factor(sample(c(rep(c(1,2,8,9),4),rep(c(3,4,6,7),8),rep(5,16))))
#summary(lme4::lmer(score ~ age + (1 | group), data = dta_grouped))
#summary(lme4::lmer(score ~ age + (1 | group0), data = dta_grouped))
```

```{r}
#| label: lm_add_residuals
pltdta_grouped <- dta_grouped
lm_naive_9 <- lm(score ~ age, data = pltdta_grouped)
pltdta_grouped$fitted_naive <- fitted(lm_naive_9)
pltdta_grouped$resid_naive <- residuals(lm_naive_9)
```

```{r}
#| label: prep_fig_resid_no_iid
#| fig-height: 5.5
#| fig-width: 5
plt_resid_naive <- ggplot(pltdta_grouped, aes(fitted_naive, resid_naive))  +
  geom_text(aes(label = group,
    colour = factor(group %in% c(1,9), 
      levels = c(FALSE, TRUE),
      labels = c("other", "1"))),
      size = 4) +
  scale_color_manual(values = c("other" = "grey70", "1" = "red"), guide = "none") +
  geom_hline(yintercept=0,lty=2,color="red") +
  labs(title="Unaccounted for Groups") + ylim(c(-4,4))
```


```{r}
#| label: lm_group_main_add_residuals
lm_group_main_9 <- lm(score ~ age + group, data = pltdta_grouped)
pltdta_grouped$fitted_group_main <- fitted(lm_group_main_9)
pltdta_grouped$resid_group_main <- residuals(lm_group_main_9)
```

```{r}
#| label: prep_fig_resid_iid
#| fig-cap: "Accounting for groups"
#| fig-height: 5.5
#| fig-width: 5
plt_resid_group <- ggplot(pltdta_grouped, aes(fitted_group_main, resid_group_main)) +
  geom_text(aes(label = group,
    colour = factor(group %in% c(1,9), 
      levels = c(FALSE, TRUE),
      labels = c("other", "1"))),
      size = 4) +
  scale_color_manual(values = c("other" = "grey70", "1" = "red"), guide = "none") +
  geom_hline(yintercept=0,lty=2,color="red") +
  labs(title="Accounted for Groups") + ylim(c(-4,4)) 
```

```{r}
#| label: fig-grouping_yes_no
#| fig-cap: "Accounting for Grouping"
#| fig-height: 5.5
#| fig-width: 10
wrap_plots(plt_resid_naive,plt_resid_group, nrow=1)
```

Note the assumption

$$
\varepsilon_i \overset{\text{i.i.d.}}{\sim} \mathcal{N}(0, \sigma^2).
$$

All residuals (variance unaccounted for by the model) assumed distributed individually and identically,  
all part of the same underlying distribution !

Any residual is fully uninformative about any other residual.
All residuals combined reflect one single distribution.

The clustering shows that grouping should be included in the model.  

## Working Example {data-toc-title="Example"}

Let's assume a SQUARE test exists

- with test results between 0 and 20
- performed on patients
- **part of one of 9 groups of patients**
- who have particular characteristics (e.g., age, ...)
- under particular conditions (e.g., treated or not)

We want to understand the conditions (e.g., treatment) under which test results are higher/lower

- **acknowledging group-members to be more similar to each-other, in average**
- by relating 
  - the data (test results ~ outcome) to 
  - the conditions (patient properties ~ predictors) 
  - &rarr; columns
- using information offered by each of the patients (replications ~ sample size) 
  - &rarr; rows

## Regression Model: grouping as Fixed variable {data-toc-title="Grouping Fixed"}

Bringing in groups

- reduces residuals
- results in residuals around 0 for each group
- removed dependence: knowledge of an instance of a group is not informative about the other instances

### Independent groups, common age effect

:::: {.columns}

Using a group identification variable to account for groups.

The regression model is the same as before, 9 groups instead of 2:

$$
y_i = \beta_0 + \beta_1 x_i + \underline{\beta_2 g_i} + \varepsilon_i,
\qquad \text{for } i = 1,\dots,n,
$$

with the usual distributional assumption:

$$
\varepsilon_i \overset{\text{i.i.d.}}{\sim} \mathcal{N}(0, \sigma^2).
$$

&nbsp;

::: {.column width="50%" .light-grey-col}

```{r}
#| label: fig-cont9_model_independent
#| fig-cap: "Grouping as Fixed Variable; Age effect considered equal over groups, high within group similarity"
#| fig-height: 5.5
#| fig-width: 10
# use group_colors
lm_group_main_9 <- lm(score ~ age + group, data = dta_grouped)
cont9_stats <- dta_grouped |>
  summarise(
    mean_score = mean(score),
    sd_score   = sd(score),
    n          = n(),
    se_mean    = sd(score) / sqrt(n()),
    .groups    = "drop"
  )

pltdta_group_main <- augment(lm_group_main_9, data = dta_grouped) |> select(-c(resid)) |>
  rename(fitted = .fitted, resid = .resid)

names(group_colors_9) <- c(1,5,9,2,6,3,7,4,8)
pred_data <- expand_grid(
  age = seq(min(dta_grouped$age), max(dta_grouped$age), length.out = 20),
  group = factor(1:9)
)
pred_data$fitted <- predict(lm_group_main_9, newdata = pred_data)

pred_data$group_gray <- factor(pred_data$group, levels = unique(pred_data$group)[order(c(1,5,9,2,6,3,7,4,8))])

plt_similarity_high <- ggplot(pltdta_group_main, aes(x = age, y = score)) +
  geom_point(aes(color = group), size = 2.5, alpha = 0.85) +
  scale_color_manual(values = group_grays_9, guide = "none") + 
  geom_segment(
    data = pltdta_group_main,
    aes(x = age, xend = age, y = score, yend = fitted),
    color = "gray60",
    alpha = 0.3,
    linewidth = 0.8
  ) +
  geom_line(
    data = pred_data,
    aes(y = fitted, color = group_gray),
    linewidth = 1.2,
    show.legend = FALSE
  ) +
  annotate(
    "text",
    x = mean(pltdta_group_main$age) ,
    y = max(pltdta_group_main$score) + .5,
    label = paste0(
      "score ≈ ",
      round(coef(lm_group_main_9)[1], 2),
      " + ",
      round(coef(lm_group_main_9)["age"], 3),
      " × age + group effect"
    ),
    size = 4.5
  ) +
  labs(
    x = "Age (continuous predictor)",
    y = "Score (outcome)",
    title = "Equal age effect for independent groups"
  ) + 
  coord_cartesian(ylim = c(10, 20)) 
plt_similarity_high
```

:::

::: {.column width="5%"}

:::

::: {.column width="45%"}

Age effect, assumed equal for each of 9 groups.  

Age on x-axis, observations on y-axis

* lines represent group specific conditional averages
  * in this case, conditional on age
  * age effect, independent of group

:::

With nine groups in the model.

```{r}
#| label: lm_grouped_main
#| echo: true
lm_group_main_9 <- lm(score ~ age + group, data = dta_grouped)
```

```{r}
#| label: tab-tidy_lm_grouped_main
tidy_gt(lm_group_main_9)
glance_gt(lm_group_main_9)
```

There is a clear age and group effect, each of the groups differs from the reference group conditional on age.  
The Mean Squared Error is huge compared to the Residual Variance.

```{r}
#| label: tab-tidy_lm_grouped_main_anova
tidy_gt(lm_group_main_9, type='anova')
```

::::

### Group specific age effect

:::: {.columns}

What if we would allow group specific age effects.

The regression model is:

$$
y_i = \beta_0 + \beta_1 \cdot x1_i + \beta_2 \cdot g2_i + \beta_3 \cdot x1_i \cdot g2_i + \varepsilon_i,
\qquad \text{for } i = 1,\dots,n,
$$

with the usual distributional assumption:

$$
\varepsilon_i \overset{\text{i.i.d.}}{\sim} \mathcal{N}(0, \sigma^2).
$$

&nbsp;

::: {.column width="50%" .light-grey-col}

```{r}
#| label: fig-cont9_model_independent_interaction
#| fig-cap: "Grouping as Fixed Variable; Age effect allowed to be different over groups, high within group similarity"
#| fig-height: 5.5
#| fig-width: 10

lm_group_int_9 <- lm(score ~ age * group, data = dta_grouped)

pltdta_group_int <- augment(lm_group_int_9, data = select(dta_grouped, -resid)) |>
  rename(fitted = .fitted, resid = .resid)

names(group_colors_9) <- c(1,5,9,2,6,3,7,4,8)
pred_data <- expand_grid(
  age = seq(min(dta_grouped$age), max(dta_grouped$age), length.out = 20),
  group = factor(1:9)
)
pred_data$fitted <- predict(lm_group_int_9, newdata = pred_data)

pred_data$group_gray <- factor(pred_data$group, levels = unique(pred_data$group)[order(c(1,5,9,2,6,3,7,4,8))])

ggplot(pltdta_group_int, aes(x = age, y = score)) +
  geom_point(aes(color = group), size = 2.5, alpha = 0.85) +
  scale_color_manual(values = group_grays_9, guide = "none") + 
  geom_segment(
    data = pltdta_group_int, 
    aes(x = age, xend = age, y = score, yend = fitted),
    color = "gray60",
    alpha = 0.3,
    linewidth = 0.8
  ) +
  geom_line(
    data = pred_data,
    aes(y = fitted, color = group_gray),
    linewidth = 1.2,
    show.legend = FALSE
  ) +
  annotate(
    "text",
    x = mean(pltdta_group_int$age) ,
    y = max(pltdta_group_int$score) + .5,
    label = paste0(
      "score ≈ ",
      round(coef(lm_group_int_9)[1], 2),
      " + ",
      round(coef(lm_group_int_9)["age"], 3),
      " × age * group effect"
    ),
    size = 4.5
  ) +
  labs(
    x = "Age (continuous predictor)",
    y = "Score (outcome)",
    title = "Different age effect for independent groups"
  ) + 
  coord_cartesian(ylim = c(10, 20)) 
```

:::

::: {.column width="5%"}

:::

::: {.column width="45%"}

The age effect is allowed to differ for each of 9 groups.  

Age on x-axis, observations on y-axis

* lines represent group specific conditional averages

:::


With nine group in the model.

```{r}
#| label: lm_grouped_interact
#| echo: true
lm_group_int_9 <- lm(score ~ age * group, data = dta_grouped)
```

```{r}
#| label: tab-tidy_lm_grouped_interact
tidy_gt(lm_group_int_9)
glance_gt(lm_group_int_9)
```

There is a clear age and group effect, each of the groups differs from the reference group conditional on age.  
The Mean Squared Error is huge compared to the Residual Variance.

```{r}
#| label: tab-tidy_lm_grouped_interact_anova
tidy_gt(lm_group_int_9, type='anova')
```

::::

## Regression Model: Why this sometimes is not enough {data-toc-title="Grouping Issues"}

:::: {.columns}

Grouping as fixed variable can be a problem,  
especially with many small clusters (groups):

- many parameters have to be estimated because many clusters
- for each cluster, parameters poorly estimated when small clusters (overfitting)
- no generalization to clusters not part of the data

::: {.column width="50%" .light-grey-col}

```{r}
#| label: fig-lack_idd
#| fig-cap: "Clustered Residuals"
#| fig-height: 5.5
#| fig-width: 7
plt_resid_naive +
  labs(title="Clustered Residuals") + 
    ylim(c(-4,4)) + 
    geom_text(aes(label = group,
      colour = factor(group %in% c(1,9), 
        levels = c(FALSE, TRUE),
        labels = c("other", "1"))),
        size = 4) +
  scale_color_manual(values = c("other" = "grey70", "1" = "red"), guide = "none")
```

:::

::: {.column width="5%"}

:::

::: {.column width="45%"}

Ignoring grouping is a problem when within group similarities exist (dependencies):

- violation of i.i.d. assumption
- **Standard errors** are too small → p-values too small → false positives inflate.
- group specific effects ignored too

:::

&nbsp;

Solution: model the within group similarities,   
once estimated the rest of the model is conditional upon it.

::::

## Note: within cluster similarity ~ between cluster difference {data-toc-title="Variance & Correlation"}

Capture dependence through estimating variance.

```{r}
#| label: fig-cont9_model_independent_interaction_0
#| fig-cap: "Grouping as Fixed Variable; low within group similarity"
#| fig-height: 5.5
#| fig-width: 10

lm_group0_main_9 <- lm(score ~ age + group0, data = dta_grouped)

pltdta_group0_main <- augment(lm_group0_main_9, data = select(dta_grouped, -resid)) |>
  rename(fitted = .fitted, resid = .resid)

names(group_colors_9) <- c(1,5,9,2,6,3,7,4,8)
pred_data <- expand_grid(
  age = seq(min(dta_grouped$age), max(dta_grouped$age), length.out = 20),
  group0 = factor(1:9)
)
pred_data$fitted <- predict(lm_group0_main_9, newdata = pred_data)

pred_data$group_gray <- factor(pred_data$group0, levels = unique(pred_data$group0)[order(c(1,5,9,2,6,3,7,4,8))])

plt_similarity_low <- ggplot(pltdta_group0_main, aes(x = age, y = score)) +
  geom_point(aes(color = group0), size = 2.5, alpha = 0.85) +
  scale_color_manual(values = group_grays_9, guide = "none") + 
  geom_segment(data = pltdta_group0_main,
    aes(x = age, xend = age, y = score, yend = fitted),
    color = "gray60",
    alpha = 0.3,
    linewidth = 0.8
  ) +
    geom_line(
  data = pred_data,
    aes(y = fitted, color = group_gray),
    linewidth = 1.2,
    show.legend = FALSE
  ) +
  annotate(
  "text",
    x = mean(pltdta_group0_main$age) ,
    y = max(pltdta_group0_main$score) + .5,
    label = paste0(
      "score ≈ ",
      round(coef(lm_group0_main_9)[1], 2),
      " + ",
      round(coef(lm_group0_main_9)["age"], 3),
      " × age + group effect"
    ),
    size = 4.5
  ) +
    labs(
    x = "Age (continuous predictor)",
    y = "Score (outcome)",
    title = "Age effect for independent groups with low within group similarity"
  ) + 
  coord_cartesian(ylim = c(10, 20)) 
```

```{r}
#| label: fig-heterogeneity_similarity
#| fig-cap: "Variance and Correlation, two sides of a coin"
#| fig-height: 5.5
#| fig-width: 10
plt_similarity_high <- plt_similarity_high  +
  labs(
    x = "Age (continuous predictor)",
    y = "Score (outcome)",
    title = "High Similarity"
  )
plt_similarity_low <- plt_similarity_low  +
  labs(
    x = "Age (continuous predictor)",
    y = "Score (outcome)",
    title = "Low Similarity"
  )
wrap_plots(plt_similarity_high,plt_similarity_low, nrow=1)
```

<!--

### Independent groups, low within group similarity

:::: {.columns}

What if groups do not differ as much.

The regression model still is:

$$
y_i = \beta_0 + \beta_1 x1_i + \beta_1 g2_i + \beta_3 x1_i * g2_i + \varepsilon_i,
\qquad \text{for } i = 1,\dots,n,
$$

with the usual distributional assumption:

$$
\varepsilon_i \overset{\text{i.i.d.}}{\sim} \mathcal{N}(0, \sigma^2).
$$

&nbsp;

::: {.column width="50%" .light-grey-col}

```{r}
#| label: fig-cont9_model_independent_interaction_0_remove
#| fig-cap: "Grouping as Fixed Variable; low within group similarity"
#| fig-height: 5.5
#| fig-width: 10

lm_group0_main_9 <- lm(score ~ age + group0, data = dta_grouped)

pltdta_group0_main <- augment(lm_group0_main_9, data = select(dta_grouped, -resid)) |>
  rename(fitted = .fitted, resid = .resid)

names(group_colors_9) <- c(1,5,9,2,6,3,7,4,8)
pred_data <- expand_grid(
  age = seq(min(dta_grouped$age), max(dta_grouped$age), length.out = 20),
  group0 = factor(1:9)
)
pred_data$fitted <- predict(lm_group0_main_9, newdata = pred_data)

pred_data$group_gray <- factor(pred_data$group0, levels = unique(pred_data$group0)[order(c(1,5,9,2,6,3,7,4,8))])

plt_similarity_low <- ggplot(pltdta_group0_main, aes(x = age, y = score)) +
  geom_point(aes(color = group0), size = 2.5, alpha = 0.85) +
  scale_color_manual(values = group_grays_9, guide = "none") + 
  geom_segment(data = pltdta_group0_main,
    aes(x = age, xend = age, y = score, yend = fitted),
    color = "gray60",
    alpha = 0.3,
    linewidth = 0.8
  ) +
    geom_line(
  data = pred_data,
    aes(y = fitted, color = group_gray),
    linewidth = 1.2,
    show.legend = FALSE
  ) +
    geom_rect(
  data = data.frame(
    xmin = max(pltdta_group0_main$age) + 0.5,
      xmax = max(pltdta_group0_main$age) + 1,
      ymin = cont9_stats$mean_score - cont9_stats$se_mean,
      ymax = cont9_stats$mean_score + cont9_stats$se_mean
    ),
    aes(xmin = xmin, xmax = xmax, ymin = ymin, ymax = ymax),
    fill = col2,
    alpha = 0.4,
    color = "black",
    linewidth = 0.5,
    inherit.aes = FALSE
  ) +
    annotate(
  "text",
    x = max(pltdta_group0_main$age) - 1.2,
    y = cont9_stats$mean_score + 0.4,
    label = "SE",
    color = col2,
    hjust = -0.1,
    size = 5
  ) +
    annotate(
  "text",
    x = mean(pltdta_group0_main$age) ,
    y = max(pltdta_group0_main$score) + .5,
    label = paste0(
      "score ≈ ",
      round(coef(lm_group0_main_9)[1], 2),
      " + ",
      round(coef(lm_group0_main_9)["age"], 3),
      " × age + group effect"
    ),
    size = 4.5
  ) +
    labs(
    x = "Age (continuous predictor)",
    y = "Score (outcome)",
    title = "Age effect for independent groups with low within group similarity"
  ) + 
  coord_cartesian(ylim = c(10, 20)) 
plt_similarity_low
```

:::

::: {.column width="5%"}

:::

::: {.column width="45%"}

The age effect is allowed to differ for each of 9 groups.  

Age on x-axis, observations on y-axis

* lines represent group specific conditional averages

:::


With nine group in the model.

```{r}
#| label: lm_weak_grouped_interact_remove
#| echo: true
lm_group0_int_9 <- lm(score ~ age * group0, data = dta_grouped)
```

```{r}
#| label: tab-tidy_lm_weak_grouped_interact_remove
tidy_gt(lm_group0_int_9)
glance_gt(lm_group0_int_9)
```

There is no evidence for an age or group effect.  
The Mean Squared Error is small compared to the Residual Variance.

```{r}
#| label: tab-tidy_lm_weak_grouped_interact_anova_rmeove
tidy_gt(lm_group0_int_9, type='anova')
```

Let's remove the interaction from the model.

```{r}
#| label: tab-lm_weak_grouped_main_remove
#| echo: true
lm_group0_main_9 <- lm(score ~ age + group0, data = dta_grouped)
```

```{r}
#| label: tab-tidy_lm_weak_grouped_main_remove
tidy_gt(lm_group0_main_9)
glance_gt(lm_group0_main_9)
```

Now there is evidence for an age effect, because it is not linked to the reference group.  
The group effect is still not present, and maybe can be removed.   
The Mean Squared Error is small compared to the Residual Variance.

```{r}
#| label: tab-tidy_lm_weak_grouped_main_anova_remove
tidy_gt(lm_group0_main_9, type='anova')
```

::::

-->

```{r}
save.image('zenzi.RData')
```
