# Additional Nesting

## Nesting Within Nesting  {data-toc-title="Higher Levels"}

Data can be nested at more than two levels: patients within wards within hospitals.

Let $y_{ijk}$ = outcome for patient $i$, in ward $j$, in hospital $k$.

::: {.equation-box}
**Level 1** (patient within ward):
$$
y_{ijk} = \pi_{0jk} + \pi_1 x_{ijk} + e_{ijk}, \qquad e_{ijk} \sim \mathcal{N}(0,\ \sigma^2)
$$

**Level 2** (ward within hospital):
$$
\pi_{0jk} = \underline{\beta_{00k}} + r_{0jk}, \qquad r_{0jk} \sim \mathcal{N}(0,\ \tau_\pi^2)
$$

**Level 3** (hospital):
$$
\underline{\beta_{00k}} = \gamma_{000} + u_{00k}, \qquad u_{00k} \sim \mathcal{N}(0,\ \tau_\beta^2)
$$

**Combined:**
$$
y_{ijk} = \underbrace{\gamma_{000} + \pi_1 x_{ijk}}_{\text{fixed}} + \underbrace{u_{00k} + r_{0jk} + e_{ijk}}_{\text{three random terms}}
$$
:::

Additional **variance components** are included to capture heterogeneity: 

- $\tau_\beta^2$ (between hospitals)
- $\tau_\pi^2$ (between wards within hospitals)
- $\sigma^2$ (within wards).

Note: different levels are assumed independent,  
within a level covariance can exist,  
between levels covariance is not possible.

```{r}
#| eval: false
#| label: fig-three-level
#| fig-cap: "Three-level structure. Each panel = one school; colours = classes within that school. Between-school differences are visible across panels; between-class differences appear within each panel."
#| fig-height: 5

set.seed(21)
n_schools3 <- 4; n_classes <- 3; n_students3 <- 20

school_u <- rnorm(n_schools3, 0, 8)
class_r  <- rnorm(n_schools3 * n_classes, 0, 4)

df_3lvl <- do.call(rbind, lapply(1:n_schools3, function(s) {
  do.call(rbind, lapply(1:n_classes, function(cl) {
    x <- runif(n_students3, 0, 10)
    y <- 50 + school_u[s] + class_r[(s-1)*n_classes + cl] +
         1.5 * x + rnorm(n_students3, 0, 3)
    data.frame(x = x, y = y,
               school = factor(s, labels = paste("School", s)),
               class  = factor((s-1)*n_classes + cl))
  }))
}))

ggplot(df_3lvl, aes(x, y, color = class)) +
  geom_point(alpha = 0.5, size = 1.8) +
  geom_smooth(method = "lm", se = FALSE, linewidth = 0.9, aes(group = class)) +
  facet_wrap(~school, nrow = 1) +
  scale_color_manual(values = rep(pal[c(1, 3, 5)], 4)) +
  labs(title = "Three-Level Structure: Students within Classes within Schools",
       subtitle = "Each panel = one school; colours = classes within that school",
       x = "SES", y = "Test score", color = "Class") +
  guides(color = "none")
```

```{r}
#| label: fit-3level
#| code-fold: false
#| eval: false
#| echo: true
m_3lvl <- lmer(y ~ x + (1 | school / class), data = df_3lvl)
```

```{r}
#| label: var-decomp
#| code-fold: false
#| eval: false
vc3      <- as.data.frame(VarCorr(m_3lvl))
vc3$pct  <- round(100 * vc3$vcov / sum(vc3$vcov), 1)
vc3[, c("grp", "vcov", "pct")]
```

## Cross-classification  {data-toc-title="Crossed"}

A level 1 unit can be associated with multiple higher level classifications *jointly*. Examples include:

- Patients measured once, classified by both **hospital** and **neighborhood**
- Ratings classified by both **rater** and **target**

In such cases, the higher-level units are **crossed** rather than nested, with a separate random effect for each classification.

Let $y_{ijk}$ denote the outcome for student $i$, in school $j$, and in neighborhood $k$.

::: {.equation-box}
**Level 1**
$$
y_{ijk} = \beta_0 + \beta_1 x_{ijk} + u_{0j} + v_{0k} + e_{ijk},
\qquad e_{ijk} \sim \mathcal{N}(0,\ \sigma^2)
$$

**Level 2 random effects**
$$
u_{0j} \sim \mathcal{N}(0,\ \tau_{\text{school}}^2),
\qquad
v_{0k} \sim \mathcal{N}(0,\ \tau_{\text{neigh}}^2)
$$

**Combined**
$$
y_{ijk} = \beta_0 + \beta_1 x_{ijk} + u_{0j} + v_{0k} + e_{ijk}
$$
:::

Additional **variance components** capture heterogeneity:

- $\tau_{\text{school}}^2$ (between hospitals)
- $\tau_{\text{neigh}}^2$ (between neighborhoods)
- $\sigma^2$ (within school–neighborhood combinations)

Here schools and neighborhoods are **independent classifications**, not nested. A patient belongs to one hospital and one neighborhood, but hospitals are not contained within neighborhoods, nor vice versa [1][2].

```{r}
#| label: fit-crossed
#| eval: false
#| echo: true
m_cc <- lmer(y ~ x + (1 | school) + (1 | neighborhood), data = df_cc)
```

Extendable to **crossed random slopes** (e.g., school-specific effects of $x$)

```{r}
#| label: fit-crossed-slopes
#| eval: false
#| echo: true
m_cc_slope <- lmer(y ~ x + (x | hospital) + (1 | neighborhood), data = df_cc)
```

```{r}
#| eval: false
#| label: fig-crossed
#| fig-cap: "Cross-classified structure. Students are classified by school and neighborhood. Schools and neighborhoods are not nested; they are crossed."
#| fig-height: 5

set.seed(22)
n_schools2 <- 4; n_neigh <- 3; n_students2 <- 15

school_u  <- rnorm(n_schools2, 0, 8)
neigh_v   <- rnorm(n_neigh, 0, 5)

df_cc <- do.call(rbind, lapply(1:n_schools2, function(s) {
  do.call(rbind, lapply(1:n_neigh, function(n) {
    x <- runif(n_students2, 0, 10)
    y <- 50 + school_u[s] + neigh_v[n] +
         1.5 * x + rnorm(n_students2, 0, 3)
    data.frame(x = x, y = y,
               school = factor(s, labels = paste("School", s)),
               neighborhood = factor(n, labels = paste("Neigh", n)))
  }))
}))

ggplot(df_cc, aes(x, y, color = school)) +
  geom_point(alpha = 0.5, size = 1.8) +
  geom_smooth(method = "lm", se = FALSE, linewidth = 0.9, aes(group = school)) +
  facet_wrap(~neighborhood, nrow = 1) +
  scale_color_manual(values = rep(pal[c(1, 3, 5)], 4)) +
  labs(title = "Cross-classified Structure: Students within Schools and Neighborhoods",
       subtitle = "Panels = neighborhoods; colours = schools (crossed, not nested)",
       x = "SES", y = "Test score", color = "School") +
  guides(color = "none")
```

```{r}
#| label: fit-cc
#| eval: false
#| echo: false
m_cc <- lmer(y ~ x + (1 | school) + (1 | neighborhood), data = df_cc)
```

```{r}
#| label: var-decomp-cc
#| eval: false
vc_cc <- as.data.frame(VarCorr(m_cc))
vc_cc$pct <- round(100 * vc_cc$vcov / sum(vc_cc$vcov), 1)
vc_cc[, c("grp", "vcov", "pct")]
```

```{r}
save.image('zenzi.RData')
```