---
title: "Workshop: Linear Mixed Models"
subtitle: "Worked example with a longitudinal treatment study"
date: today
published-title: "Rendered"
callout-appearance: simple
callout-icon: false
format:
  html:
    number-depth: 3
execute:
  echo: true
  warning: false
  message: false
---

```{r}
#| label: setup
library(tidyverse)
library(lme4)
library(lmerTest)
library(broom.mixed)
library(gt)
library(ggplot2)
library(patchwork)

# Utility: gt-wrapped tidy output
tidy_gt <- function(model, ...) {
  broom.mixed::tidy(model, ...) |>
    mutate(across(where(is.numeric), \(x) round(x, 4))) |>
    gt() |>
    tab_options(table.font.size = 13)
}

glance_gt <- function(model) {
  broom.mixed::glance(model) |>
    mutate(across(where(is.numeric), \(x) round(x, 3))) |>
    gt() |>
    tab_options(table.font.size = 13)
}

col_patient  <- "#003399"
col_pop      <- "#e31a1c"
col_accent   <- "#ff6600"
col_grey     <- "grey60"
```

## Mixed Models: exercises {.unnumbered}

### Setup {data-toc-title="Example"}

#### Data

We use a **longitudinal treatment study** with repeated measurements on the same patients over time.

The dataset contains:

- repeated outcomes for each patient over visits
- a **treatment group** indicator (Treatment `T` vs Control `C`)
- a continuous baseline covariate: **age**
- slightly **unbalanced follow-up**: each patient has between 3 and 5 visits
- patient-specific heterogeneity in both baseline and rate of change

This means the dataset is very close to a typical clinical repeated-measures setting, but still rich enough to motivate random slopes, model comparison, prediction, and diagnostics.

```{r}
#| label: load-data
dta_long <- read_csv("longitudinal_treatment_study.csv", show_col_types = FALSE) |>
  mutate(
    patient_id = factor(patient_id),
    treatment = factor(treatment, levels = c("C", "T")),
    visit_f = factor(visit),
    age_c = age - mean(age),
    age_z = as.numeric(scale(age))
  )

glimpse(dta_long)
```

```{r}
#| label: tab-first-rows
dta_long |> slice(1:12) |> gt()
```

#### Aim and Method

We want to understand how the score evolves over time and whether treatment matters, while accounting for the fact that repeated measurements on the same patient are correlated.

We use **linear mixed models (LMMs)** to decompose variation into:

- **fixed effects**: population-level mean trends over visit, treatment, and age
- **random effects**: patient-specific deviations from the average trajectory
- **residual variance**: remaining within-patient noise

### Wide vs. Long Format: Exercise {data-toc-title="Data Reshaping"}

Clinical datasets are often delivered in **wide format**, with one row per patient and a separate column for each visit. Mixed models, however, require **long format**, with one row per observation.

```{r}
#| label: wide-format
dta_wide <- dta_long |>
  select(patient_id, treatment, age, visit, score) |>
  pivot_wider(
    id_cols = c(patient_id, treatment, age),
    names_from = visit,
    values_from = score,
    names_prefix = "visit_"
  )

dta_wide |> slice(1:8) |> gt()
```

Patients with fewer than 5 visits have `NA` values in later visit columns. In long format those rows simply do not exist, which is more natural for modelling unbalanced longitudinal data.

```{r}
#| label: long-format
dta_long_from_wide <- dta_wide |>
  pivot_longer(
    cols = starts_with("visit_"),
    names_to = "visit",
    names_prefix = "visit_",
    values_to = "score"
  ) |>
  filter(!is.na(score)) |>
  mutate(visit = as.integer(visit)) |>
  arrange(patient_id, visit)

dta_long_from_wide |> slice(1:12) |> gt()
```

::: {.callout-warning title="Exercise"}
Repeat the wide-to-long transformation yourself.

- Why is long format more natural for mixed models?
- How does long format handle imbalance more elegantly than wide format?
:::

### Exploring the Data: Exercise {data-toc-title="Exploration"}

Before fitting models, inspect the data for anomalies, imbalance, and patterns that matter for modelling.

#### Descriptive checks

```{r}
#| label: desc-counts
dta_long |>
  count(treatment)
```

```{r}
#| label: desc-visits
dta_long |>
  count(patient_id, name = "n_visits") |>
  count(n_visits)
```

```{r}
#| label: desc-age
dta_long |>
  distinct(patient_id, treatment, age) |>
  ggplot(aes(x = age, fill = treatment)) +
  geom_histogram(bins = 12, alpha = 0.65, position = "identity") +
  scale_fill_manual(values = c(C = col_patient, T = col_accent)) +
  labs(x = "Age", y = "Count", title = "Age distribution by treatment") +
  theme_bw(base_size = 13)
```

#### Trajectories

A spaghetti plot shows individual longitudinal profiles and the mean trend per treatment group.

```{r}
#| label: fig-spaghetti
#| fig-cap: "Individual trajectories by treatment group"
#| fig-height: 5
#| fig-width: 10
dta_long |>
  ggplot(aes(x = visit, y = score, group = patient_id, colour = treatment)) +
  geom_line(alpha = 0.45, linewidth = 0.7) +
  geom_point(alpha = 0.65, size = 1.5) +
  stat_summary(
    aes(group = treatment),
    fun = mean, geom = "line",
    linewidth = 1.8, linetype = "dashed"
  ) +
  scale_colour_manual(values = c(C = col_patient, T = col_accent)) +
  facet_wrap(~ treatment, labeller = as_labeller(c(C = "Control", T = "Treatment"))) +
  labs(
    x = "Visit", y = "Score",
    title = "Individual trajectories per patient",
    subtitle = "Dashed line = treatment-specific mean trajectory"
  ) +
  theme_bw(base_size = 13) +
  theme(legend.position = "none")
```

To reduce clutter, one can show only the most and least variable trajectories within each treatment group.

```{r}
#| label: fig-spaghetti-var
#| fig-cap: "Most and least variable trajectories by treatment group"
#| fig-height: 5
#| fig-width: 6
sel_patients <- dta_long |>
  group_by(treatment, patient_id) |>
  summarise(within_var = var(score, na.rm = TRUE), .groups = "drop") |>
  group_by(treatment) |>
  slice_min(within_var, n = 1, with_ties = FALSE) |>
  bind_rows(
    dta_long |>
      group_by(treatment, patient_id) |>
      summarise(within_var = var(score, na.rm = TRUE), .groups = "drop") |>
      group_by(treatment) |>
      slice_max(within_var, n = 1, with_ties = FALSE)
  ) |>
  distinct(treatment, patient_id)

dta_long |>
  semi_join(sel_patients, by = c("treatment", "patient_id")) |>
  ggplot(aes(x = visit, y = score, group = patient_id, colour = treatment)) +
  geom_line(alpha = 0.6, linewidth = 0.8) +
  geom_point(alpha = 0.8, size = 1.6) +
  scale_colour_manual(values = c(C = col_patient, T = col_accent)) +
  labs(
    x = "Visit", y = "Score",
    title = "Most and least variable trajectories"
  ) +
  theme_bw(base_size = 13) +
  theme(legend.position = "none")
```

#### Within- vs. Between-patient variation

A key concept is the distinction between variation **between** patients and variation **within** patients over time.

```{r}
#| label: fig-within-between
#| fig-cap: "Within vs. between-patient variation"
#| fig-height: 5.5
patient_means <- dta_long |>
  group_by(patient_id, treatment) |>
  summarise(mean_score = mean(score), .groups = "drop")

p_between <- patient_means |>
  ggplot(aes(x = reorder(patient_id, mean_score), y = mean_score, fill = treatment)) +
  geom_col(width = 0.7, alpha = 0.85) +
  scale_fill_manual(values = c(C = col_patient, T = col_accent)) +
  labs(
    x = "Patient (sorted by mean)", y = "Mean score",
    title = "Between-patient variation",
    subtitle = "Each bar is a patient's mean over visits"
  ) +
  theme_bw(base_size = 12) +
  theme(axis.text.x = element_blank(), axis.ticks.x = element_blank())

p_within <- dta_long |>
  left_join(patient_means, by = c("patient_id", "treatment")) |>
  mutate(within_dev = score - mean_score) |>
  ggplot(aes(x = factor(visit), y = within_dev, fill = treatment)) +
  geom_boxplot(alpha = 0.75) +
  geom_hline(yintercept = 0, linetype = "dashed", colour = col_grey) +
  scale_fill_manual(values = c(C = col_patient, T = col_accent)) +
  labs(
    x = "Visit", y = "Score - patient mean",
    title = "Within-patient variation",
    subtitle = "Deviation from each patient's own average"
  ) +
  theme_bw(base_size = 12)

wrap_plots(p_between, p_within, nrow = 1)
```

::: {.callout-warning title="Exercise"}
Explain which part of the variation could be captured by:

- a random intercept
- a random slope for visit
- the residual variance
:::

### Estimating Mixed Models {data-toc-title="Estimation"}

#### Model recap

We begin with a random-intercept model:

$$
\text{score}_{ij} = \beta_0 + \beta_1\,\text{visit}_{ij} + \beta_2\,\text{treatment}_i + \beta_3\,\text{age}_{ci} + u_{0i} + \varepsilon_{ij}
$$

where:

$$
u_{0i} \sim \mathcal{N}(0, \sigma^2_{u_0}), \qquad
\varepsilon_{ij} \sim \mathcal{N}(0, \sigma^2), \qquad
u_{0i} \perp \varepsilon_{ij}.
$$

The random intercept captures a patient's baseline deviation from the population mean.

#### Fitting models

##### Model 0 — Intercept only

```{r}
#| label: mod0
mod0 <- lmer(score ~ 1 + (1 | patient_id), data = dta_long, REML = TRUE)
summary(mod0)
```

##### Model 1a — Add continuous time

```{r}
#| label: mod1
mod1 <- lmer(score ~ visit + (1 | patient_id), data = dta_long, REML = FALSE)
tidy_gt(mod1, effects = "fixed")
```

##### Model 1b — Add categorical time

```{r}
#| label: mod1b
mod1b <- lmer(score ~ visit_f + (1 | patient_id), data = dta_long, REML = FALSE)
tidy_gt(mod1b, effects = "fixed")
```

##### Model 2 — Add treatment

```{r}
#| label: mod2
mod2 <- lmer(score ~ visit + treatment + (1 | patient_id), data = dta_long, REML = FALSE)
tidy_gt(mod2, effects = "fixed")
```

##### Model 3 — Add centered age

```{r}
#| label: mod3
mod3 <- lmer(score ~ visit + treatment + age_c + (1 | patient_id), data = dta_long, REML = FALSE)
tidy_gt(mod3, effects = "fixed")
```

Centering changes the interpretation of the intercept: it is now the expected score at `visit = 0` for the reference treatment group at the mean age.

##### Model 3b — Add treatment-by-visit interaction

```{r}
#| label: mod3b
mod3b <- lmer(score ~ visit * treatment + age_c + (1 | patient_id), data = dta_long, REML = FALSE)
tidy_gt(mod3b, effects = "fixed")
```

This interaction allows the treatment groups to evolve differently over time.

##### Model 4 — Add random slope for time

```{r}
#| label: mod4
mod4 <- lmer(score ~ visit + treatment + age_c + (1 + visit | patient_id),
             data = dta_long, REML = FALSE)
tidy_gt(mod4, effects = "fixed")
```

##### Model 5 — Add interaction and random slope

```{r}
#| label: mod5
mod5 <- lmer(score ~ visit * treatment + age_c + (1 + visit | patient_id),
             data = dta_long, REML = FALSE)
tidy_gt(mod5, effects = "fixed")
```

::: {.callout-warning title="Exercise"}
Interpret:

- the coefficient for `visit`
- the coefficient for `treatmentT`
- the coefficient for `age_c`
- the interaction `visit:treatmentT`
:::

### Model Comparison {data-toc-title="Model Comparison"}

Use likelihood ratio tests for nested model comparisons. Compare fixed effects under **ML**.

#### Evaluation of fixed effects

```{r}
#| label: anova-fixed
anova(mod1, mod2, mod3, mod3b)
```

This lets you assess whether treatment, age, and the treatment-by-visit interaction improve model fit.

#### Evaluation of random effects

```{r}
#| label: anova-random
anova(mod3, mod4)
```

```{r}
#| label: anova-random-interaction
anova(mod3b, mod5)
```

Testing a variance component against 0 places the null on the boundary of the parameter space, so the naive LRT p-value is approximate.

For teaching purposes, it is still useful, but one should note that a mixture distribution or a parametric bootstrap is preferable in more formal settings.

#### AIC / BIC summary table

```{r}
#| label: model-comparison-table
model_list <- list(
  "M0: intercept only" = mod0,
  "M1: + visit" = mod1,
  "M1b: visit categorical" = mod1b,
  "M2: + treatment" = mod2,
  "M3: + age" = mod3,
  "M3b: + visit:treatment" = mod3b,
  "M4: + random slope" = mod4,
  "M5: + interaction + random slope" = mod5
)

map_dfr(model_list, broom.mixed::glance, .id = "model") |>
  select(model, AIC, BIC, logLik, df.residual) |>
  mutate(across(where(is.numeric), \(x) round(x, 1))) |>
  gt() |>
  tab_style(
    style = cell_fill(color = "#d4edda"),
    locations = cells_body(rows = AIC == min(AIC))
  )
```

::: {.callout-warning title="Exercise"}
Do the AIC/BIC conclusions support the LRT-based conclusions?

Would you retain the interaction term? Would you retain the random slope?
:::

### Fixed and Random Effect Estimates {data-toc-title="Estimates"}

Suppose we select `mod4` as the final working model: linear visit effect, treatment, centered age, and a random slope for visit.

We now re-fit it using **REML** for variance estimation.

```{r}
#| label: final-model
mod_final <- lmer(score ~ visit + treatment + age_c + (1 + visit | patient_id),
                  data = dta_long, REML = TRUE)
summary(mod_final)
```

#### Fixed effects

```{r}
#| label: tab-fixed-effects
tidy_gt(mod_final, effects = "fixed", conf.int = TRUE)
```

#### Variance components

```{r}
#| label: tab-random-effects
tidy_gt(mod_final, effects = "ran_pars")
```

#### Intraclass Correlation Coefficient (ICC)

For a random-intercept model, the ICC is:

$$
ICC = \frac{\sigma^2_{u_0}}{\sigma^2_{u_0} + \sigma^2}
$$

Here, because the final model contains a random slope as well, the simple ICC is mainly a rough descriptive summary rather than a complete description of the covariance structure.

```{r}
#| label: icc
mod_icc <- lmer(score ~ visit + treatment + age_c + (1 | patient_id),
                data = dta_long, REML = TRUE)

vc <- as.data.frame(VarCorr(mod_icc))
sigma2_u0  <- vc$vcov[vc$grp == "patient_id"]
sigma2_eps <- vc$vcov[vc$grp == "Residual"]
icc <- sigma2_u0 / (sigma2_u0 + sigma2_eps)

tibble(
  sigma2_u0 = round(sigma2_u0, 3),
  sigma2_eps = round(sigma2_eps, 3),
  ICC = round(icc, 3)
) |>
  gt()
```

::: {.callout-warning title="Exercise"}
Why is the ICC easiest to interpret in a random-intercept model?

How does that interpretation become more subtle once random slopes are added?
:::

### Prediction

Predictions can be made for:

- patients already observed in the data
- a population-average trajectory
- a new patient not observed before

#### Population-Level Predictions {data-toc-title="Population Predictions"}

Population-level predictions use only the fixed effects.

```{r}
#| label: fig-population-predictions
#| fig-cap: "Population-level predicted trajectories by treatment"
#| fig-height: 4.5
new_pop <- expand_grid(
  visit = 1:5,
  treatment = c("C", "T"),
  age_c = 0,
  patient_id = NA
) |>
  mutate(
    treatment = factor(treatment, levels = c("C", "T")),
    pred_pop = predict(mod_final, newdata = cur_data(), re.form = NA, allow.new.levels = TRUE)
  )

ggplot(new_pop, aes(x = visit, y = pred_pop, colour = treatment, group = treatment)) +
  geom_line(linewidth = 1.6) +
  geom_point(size = 3) +
  scale_colour_manual(values = c(C = col_patient, T = col_accent)) +
  labs(
    x = "Visit", y = "Predicted score",
    title = "Population-level predicted trajectories",
    subtitle = "At mean age; random effects set to zero"
  ) +
  theme_bw(base_size = 13)
```

#### Subject-Specific Predictions {data-toc-title="Subject Predictions"}

Subject-specific predictions condition on the estimated random effects.

```{r}
#| label: fig-subject-predictions
#| fig-cap: "Subject-specific predicted vs observed trajectories"
#| fig-height: 6
dta_pred <- dta_long |>
  mutate(
    pred_pop = predict(mod_final, re.form = NA),
    pred_subj = predict(mod_final)
  )

dta_pred |>
  filter(patient_id %in% levels(patient_id)[1:12]) |>
  ggplot(aes(x = visit)) +
  geom_point(aes(y = score), colour = "grey40", size = 2) +
  geom_line(aes(y = pred_subj, colour = treatment), linewidth = 1.1) +
  geom_line(aes(y = pred_pop, colour = treatment), linewidth = 0.7,
            linetype = "dashed", alpha = 0.7) +
  scale_colour_manual(values = c(C = col_patient, T = col_accent)) +
  facet_wrap(~ patient_id, ncol = 4) +
  labs(
    x = "Visit", y = "Score",
    title = "Subject-specific (solid) vs population (dashed) predictions",
    subtitle = "Observed scores shown as points"
  ) +
  theme_bw(base_size = 11) +
  theme(legend.position = "top")
```

#### Predicting a New Patient {data-toc-title="New Patient"}

For a new patient, there is no estimated patient-specific random effect. The point prediction is therefore population-level.

```{r}
#| label: new-patient-prediction
new_patient <- tibble(
  visit = 1:5,
  treatment = factor("T", levels = c("C", "T")),
  age_c = 40 - mean(dta_long$age),
  patient_id = NA
) |>
  mutate(
    pred = predict(mod_final, newdata = cur_data(), re.form = NA,
                   allow.new.levels = TRUE)
  )

new_patient |>
  mutate(pred = round(pred, 2)) |>
  gt() |>
  tab_header(
    title = "Predicted scores for a new 40-year-old treatment patient",
    subtitle = "Population-level prediction: random effects set to zero"
  )
```

::: {.callout-warning title="Exercise"}
Explain why the point prediction for a new patient equals the population-level prediction.

Why is the uncertainty for a new patient still larger than uncertainty for the population mean?
:::

### Diagnostics

Good diagnostics evaluate whether the model is a reasonable approximation to the data.

```{r}
#| label: diagnostic-setup
res_df <- dta_long |>
  mutate(
    fitted_val = fitted(mod_final),
    resid_val = resid(mod_final),
    std_resid = resid_val / sigma(mod_final)
  )

blups <- ranef(mod_final)$patient_id |>
  rownames_to_column("patient_id") |>
  rename(u0_hat = `(Intercept)`, u1_hat = visit)
```

#### Residuals vs. Fitted

```{r}
#| label: fig-resid-fitted
#| fig-cap: "Residuals vs fitted values"
#| fig-height: 4
res_df |>
  ggplot(aes(x = fitted_val, y = resid_val)) +
  geom_point(alpha = 0.6, colour = col_patient) +
  geom_hline(yintercept = 0, linetype = "dashed", colour = col_grey) +
  geom_smooth(se = FALSE, colour = col_pop, linewidth = 1) +
  labs(x = "Fitted values", y = "Residuals", title = "Residuals vs Fitted") +
  theme_bw(base_size = 13)
```

#### Q-Q plot of residuals

```{r}
#| label: fig-qq-resid
#| fig-cap: "Q-Q plot of standardised residuals"
#| fig-height: 4
res_df |>
  ggplot(aes(sample = std_resid)) +
  stat_qq(alpha = 0.6, colour = col_patient) +
  stat_qq_line(colour = col_pop, linewidth = 1) +
  labs(
    title = "Normal Q-Q: Standardised Residuals",
    x = "Theoretical Quantiles", y = "Sample Quantiles"
  ) +
  theme_bw(base_size = 13)
```

#### Q-Q plot of random effects

```{r}
#| label: fig-qq-blups
#| fig-cap: "Q-Q plots of estimated random effects"
#| fig-height: 4
p_u0 <- blups |>
  ggplot(aes(sample = u0_hat)) +
  stat_qq(alpha = 0.7, colour = col_accent) +
  stat_qq_line(colour = col_pop, linewidth = 1) +
  labs(title = "Random intercepts", x = "Theoretical Quantiles", y = "Sample Quantiles") +
  theme_bw(base_size = 12)

p_u1 <- blups |>
  ggplot(aes(sample = u1_hat)) +
  stat_qq(alpha = 0.7, colour = col_patient) +
  stat_qq_line(colour = col_pop, linewidth = 1) +
  labs(title = "Random slopes", x = "Theoretical Quantiles", y = "Sample Quantiles") +
  theme_bw(base_size = 12)

p_u0 + p_u1
```

#### Scale-Location plot

```{r}
#| label: fig-scale-location
#| fig-cap: "Scale-location plot"
#| fig-height: 4
res_df |>
  mutate(sqrt_abs_resid = sqrt(abs(std_resid))) |>
  ggplot(aes(x = fitted_val, y = sqrt_abs_resid)) +
  geom_point(alpha = 0.6, colour = col_patient) +
  geom_smooth(se = FALSE, colour = col_pop, linewidth = 1) +
  labs(
    x = "Fitted values",
    y = expression(sqrt("|Standardised residuals|")),
    title = "Scale-Location"
  ) +
  theme_bw(base_size = 13)
```

#### Caterpillar plot of random intercepts

```{r}
#| label: fig-caterpillar-int
#| fig-cap: "Caterpillar plot of random intercepts"
#| fig-height: 5.5
re_condvar <- ranef(mod_final, condVar = TRUE)$patient_id
postvar <- attr(re_condvar, "postVar")

blup_int <- tibble(
  patient_id = rownames(re_condvar),
  u0_hat = re_condvar[, "(Intercept)"],
  se = sqrt(postvar[1, 1, ])
) |>
  mutate(
    lower = u0_hat - 1.96 * se,
    upper = u0_hat + 1.96 * se
  ) |>
  arrange(u0_hat) |>
  mutate(patient_id = factor(patient_id, levels = patient_id))

blup_int |>
  ggplot(aes(x = patient_id, y = u0_hat)) +
  geom_hline(yintercept = 0, linetype = "dashed", colour = col_grey) +
  geom_errorbar(aes(ymin = lower, ymax = upper), width = 0.3, colour = col_grey) +
  geom_point(size = 2.4, colour = col_patient) +
  labs(
    x = "Patient ID (sorted)", y = expression(hat(u)[i]),
    title = "Caterpillar plot: estimated random intercepts"
  ) +
  theme_bw(base_size = 12) +
  theme(axis.text.x = element_text(angle = 90, hjust = 1, size = 7))
```

#### Caterpillar plot of random slopes

```{r}
#| label: fig-caterpillar-slope
#| fig-cap: "Caterpillar plot of random slopes"
#| fig-height: 5.5
blup_slope <- tibble(
  patient_id = rownames(re_condvar),
  u1_hat = re_condvar[, "visit"],
  se = sqrt(postvar[2, 2, ])
) |>
  mutate(
    lower = u1_hat - 1.96 * se,
    upper = u1_hat + 1.96 * se
  ) |>
  arrange(u1_hat) |>
  mutate(patient_id = factor(patient_id, levels = patient_id))

blup_slope |>
  ggplot(aes(x = patient_id, y = u1_hat)) +
  geom_hline(yintercept = 0, linetype = "dashed", colour = col_grey) +
  geom_errorbar(aes(ymin = lower, ymax = upper), width = 0.3, colour = col_grey) +
  geom_point(size = 2.4, colour = col_accent) +
  labs(
    x = "Patient ID (sorted)", y = expression(hat(u)[i]),
    title = "Caterpillar plot: estimated random slopes"
  ) +
  theme_bw(base_size = 12) +
  theme(axis.text.x = element_text(angle = 90, hjust = 1, size = 7))
```

::: {.callout-warning title="Exercise"}
Review the diagnostics and discuss:

- evidence for non-linearity
- evidence for heteroscedasticity
- departures from normality
- whether some patients appear unusually influential
:::

### Possible extensions

This dataset also allows several additional exercises:

- compare linear versus categorical visit effects
- evaluate a treatment-by-visit interaction
- compare centered age versus standardized age
- discuss the ICC in random-intercept versus random-slope models
- compare subject-specific and population-level predictions
- discuss prediction for unseen patients
- inspect whether a simpler random-intercept model would be adequate

These extensions make it a good bridge between a toy simulated example and more realistic applied longitudinal data.