---
title: "Workshop: Linear Mixed Models"
subtitle: "Getting our hands dirty"
date: today
published-title: "Rendered"
callout-appearance: simple
callout-icon: false
format:
  html:
    number-depth: 3
---

```{r}
#| label: setup
#| echo: false

library(tidyverse)
library(lme4)
library(lmerTest)
library(broom.mixed)
library(gt)
library(ggplot2)
library(patchwork)
library(writexl)

# Utility: gt-wrapped tidy output
tidy_gt <- function(model, ...) {
  broom.mixed::tidy(model, ...) |>
    mutate(across(where(is.numeric), \(x) round(x, 4))) |>
    gt() |>
    tab_options(table.font.size = 13)
}

glance_gt <- function(model) {
  broom.mixed::glance(model) |>
    mutate(across(where(is.numeric), \(x) round(x, 3))) |>
    gt() |>
    tab_options(table.font.size = 13)
}

# Colour palette
col_patient  <- "#003399"
col_pop      <- "#e31a1c"
col_accent   <- "#ff6600"
col_grey     <- "grey60"
```

## Mixed Models: exercises {.unnumbered}

### Setup {data-toc-title="Example"}

#### Data

Suppose a **SQUARE** test exists:

- performed *repeatedly* on the same patients over time
- each patient has between 3 and 5 visits (slightly unbalanced)
- patients differ by **treatment group** (Treatment T vs Control C) and **baseline age**
- the test returns a score between 0 and 30

The data for this working example is simulated to have 30 patients, 3–5 visits each.

```{r}
#| label: simulate-data
#| echo: false

set.seed(42)

n_patients <- 30
max_visits  <- 5

# Fixed-effect parameters
beta_0        <- 15    # grand intercept
beta_time     <- 1.2   # score improves ~1.2 pts per visit
beta_trt      <- 2.5   # treatment B adds 2.5 pts
beta_age      <- -0.15 # older patients score slightly lower
sigma_u0      <- 3.2   # SD of patient-specific intercept
sigma_residual <- 2.0  # residual SD

patient_info <- tibble(
  patient_id  = 1:n_patients,
  treatment   = rep(c("T", "C"), each = n_patients / 2),
  age         = round(runif(n_patients, 25, 65)),
  n_visits    = sample(3:5, n_patients, replace = TRUE),
  u0          = rnorm(n_patients, 0, sigma_u0)   # random intercept
)

# Build long-form data
dta_long <- patient_info |>
  rowwise() |>
  reframe(
patient_id = patient_id,
    treatment  = treatment,
    age        = age,
    u0         = u0,
    visit      = 1:n_visits
  ) |>
    mutate(
  trt_B   = as.integer(treatment == "B"),
    mu      = beta_0 + beta_time * visit + beta_trt * trt_B +
      beta_age * (age - mean(patient_info$age)) + u0,
    score   = round(mu + rnorm(n(), 0, sigma_residual), 1),
    score   = pmax(0, pmin(30, score))           # clamp to [0, 30]
  ) |>
    select(patient_id, treatment, age, visit, score)
  
  glimpse(dta_long)
```

#### Aim and Method

We want to understand how the score evolves over time and whether treatment matters — while **properly accounting for the fact that repeated measurements on the same patient are correlated**.

We use a **Linear Mixed Models (LMM)**, which decompose variance into:

- **Fixed effects** — population-level trends (time, treatment, age)
- **Random effects** — patient-specific deviations from those trends
- **Residual variance** — unexplained within-patient noise

### Wide vs. Long Format: Exercise {data-toc-title="Data Reshaping"}

Clinical datasets are often delivered in **wide format** — one row per patient, separate columns for each visit. This is used for multivariate analysis like for example repeated measures ANOVA. 

Mixed models require **long format** (univariate) — one row per observation, used for univariate analysis.

```{r}
#| label: wide-format
#| echo: false
dta_wide <- dta_long |>
  pivot_wider(
    id_cols     = c(patient_id, treatment, age),
    names_from  = visit,
    values_from = score,
    names_prefix = "visit_"
  )

dta_wide |> slice(1:8) |> gt()
```

The simulated long-form dataset and its wide-form representation are exported to Excel, with each assigned a tab.

```{r}
write_xlsx(
  list(
    patients = dta_wide,
    scores = dta_long
  ),
  path = "ws-mixed-data.xlsx"
)
save(dta_long, dta_wide, file="ws-mixed-data.RData")
```

Patients with fewer than 5 visits get `NA` in the later visit columns. In long format, those rows simply don't exist — much cleaner for modelling. Patient specific timepoints are also easily dealt with in long form.

::: {.callout-warning title="SPSS/Excel: Reshape: wide → long"}
After showing a reshape in Excel and SPSS,  
the challenge is to repeat it for the simulated data.

Notice how a univariate representation more elegantly deals with imbalance.  
:::

```{r}
#| label: long-format
dta_long_from_wide <- dta_wide |>
  pivot_longer(
    cols         = starts_with("visit_"),
    names_to     = "visit",
    names_prefix = "visit_",
    values_to    = "score"
  ) |>
  filter(!is.na(score)) |>
  mutate(visit = as.integer(visit)) |>
  arrange(patient_id, visit)

dta_long_from_wide |> slice(1:12) |> gt()
```

### Exploring the Data: Exercise {data-toc-title="Exploration"}

Before modeling and testing,  
look at your data to ensure all is as expected, and detect anomalies.

- unexpected skewness
- impossible values
- sparse categories
- unexpected imbalance across groups
- ... 

which require some thought first 

- transformations
- methods
- stratification
- ...

Visualisations focus on key properties, part of your story 

- of the research question, assumptions, ...
- respecting the nature of the data
- clearly, without clutter

#### Trajectories

For the analysis of longitudinal data, show the trajectories with a spaghetti plot.

```{r}
#| label: fig-spaghetti
#| fig-cap: "Individual trajectories by treatment group"
#| fig-height: 5
#| fig-width: 10

dta_long |>
  ggplot(aes(x = visit, y = score, group = patient_id, colour = treatment)) +
  geom_line(alpha = 0.5, linewidth = 0.7) +
  geom_point(alpha = 0.7, size = 1.5) +
  stat_summary(
    aes(group = treatment),
    fun = mean, geom = "line",
    linewidth = 1.8, linetype = "dashed"
  ) +
  scale_colour_manual(values = c(C = col_patient, T = col_accent)) +
  facet_wrap(~ treatment, labeller = as_labeller(c(
  C = "Control",
  T = "Treatment"))) +
  labs(
    x = "Visit", y = "SQUARE test score",
    title = "Individual trajectories per patient",
    subtitle = "Dashed line = group mean trajectory"
  ) +
  theme_bw(base_size = 13) +
  theme(legend.position = "none")
```

To avoid clutter:

- facet, color, ...
- subset random, or guided by relevant properties (mean, variance, ...)


```{r}
#| label: fig-spaghetti-var
#| fig-cap: "Most/Least variable trajectories by treatment group"
#| fig-height: 5
#| fig-width: 6

sel_patients <- dta_long |>
  group_by(treatment, patient_id) |>
  summarise(within_var = var(score, na.rm = TRUE), .groups = "drop") |>
  group_by(treatment) |>
  slice_min(within_var, n = 1, with_ties = FALSE) |>
  bind_rows(
    dta_long |>
      group_by(treatment, patient_id) |>
      summarise(within_var = var(score, na.rm = TRUE), .groups = "drop") |>
      group_by(treatment) |>
      slice_max(within_var, n = 1, with_ties = FALSE)
  ) |>
  distinct(treatment, patient_id)

dta_plot <- dta_long |>
  semi_join(sel_patients, by = c("treatment", "patient_id"))

dta_plot |>
  ggplot(aes(x = visit, y = score, group = patient_id, colour = treatment)) +
  geom_line(alpha = 0.5, linewidth = 0.7) +
  geom_point(alpha = 0.7, size = 1.5) +
  scale_colour_manual(
    values = c(C = col_patient, T = col_accent),
    labels = c(C = "Control", T = "Treatment")
  ) +
  labs(
    x = "Visit", y = "SQUARE test score",
    title = "Most/least variable trajectories"
  ) +
  theme_bw(base_size = 13) +
  theme(legend.position = "none")
```


::: {.callout-warning title="SPSS: Spaghetti plots"}
After showing how to show individual trajectories in SPSS,  
the challenge is to repeat it for the simulated data.

And pretending that there are too many to show, how would you select a subset ?  
:::

#### Within- vs. Between-patient variation

A key concept in mixed models is the distinction between **within-patient** (longitudinal) and **between-patient** (cross-sectional) variation.

Possible visualisations of the heterogeneity over patients and the remaining residual variance, for example per time point are given for completeness:

```{r}
#| label: fig-within-between
#| fig-cap: "Within vs. between-patient variation"
#| fig-height: 5.5

patient_means <- dta_long |>
  group_by(patient_id, treatment) |>
  summarise(mean_score = mean(score), .groups = "drop")

p_between <- patient_means |>
  ggplot(aes(x = reorder(patient_id, mean_score), y = mean_score,
             fill = treatment)) +
  geom_col(width = 0.7, alpha = 0.8) +
  scale_fill_manual(values = c(C = col_patient, T = col_accent)) +
  labs(x = "Patient (sorted by mean)", y = "Mean score",
       title = "Between-patient variation",
       subtitle = "Each bar = one patient's average across visits") +
  theme_bw(base_size = 12) +
  theme(axis.text.x = element_blank(), axis.ticks.x = element_blank())

p_within <- dta_long |>
  left_join(patient_means, by = c("patient_id", "treatment")) |>
  mutate(within_dev = score - mean_score) |>
  ggplot(aes(x = factor(visit), y = within_dev, fill = treatment)) +
  geom_boxplot(alpha = 0.7) +
  geom_hline(yintercept = 0, linetype = "dashed", colour = col_grey) +
  scale_fill_manual(values = c(C = col_patient, T = col_accent)) +
  labs(x = "Visit", y = "Score − patient mean",
       title = "Within-patient variation",
       subtitle = "Deviation from each patient's own average") +
  theme_bw(base_size = 12)

#p_between / p_within
```

```{r}
#| label: fig-grouping-yes-no
#| fig-cap: "Accounting for Grouping"
#| fig-height: 5.5
#| fig-width: 10
wrap_plots(p_between,p_within, nrow=1)
```

The left plot shows how patient averages differ, reflecting heterogeneity, also linking it to treatment. This could be captures by a random intercept.

The right plot offers a look at the fixed effects, time and treatment, as well as the within patient variability.


::: {.callout-warning title="SPSS: Variance within and between"}
After showing a few plots highlighting between and within variability,  
the challenge is to repeat it for the simulated data.

Would you do it differently, why ?
:::

### Estimating Mixed Models {data-toc-title="Estimation"}

#### The Model: recap

We specify a **random-intercept model**:

$$
\text{score}_{ij} = \underbrace{\beta_0 + \beta_1\,\text{visit}_{ij} + \beta_2\,\text{trt}_{i} + \beta_3\,\text{age}_{i}}_{\text{fixed effects}} + \underbrace{u_{0i}}_{\text{random intercept}} + \underbrace{\varepsilon_{ij}}_{\text{residual}},
$$

for patient $i = 1,\dots,30$ and visit $j = 1,\dots,n_i$, where:

$$
u_{0i} \overset{\text{i.i.d.}}{\sim} \mathcal{N}(0,\, \sigma^2_{u_0}), \qquad
\varepsilon_{ij} \overset{\text{i.i.d.}}{\sim} \mathcal{N}(0,\, \sigma^2),
$$

and $u_{0i} \perp \varepsilon_{ij}$.

The random intercept $u_{0i}$ captures each patient's **systematic deviation** from the population mean — their personal "baseline offset".

#### Fitting Models

Build up models, step by step.

##### Model 0 — Intercept only (pooled)

Variance decomposition: within and between patients.  
Ignores repeated-measures structure, and fixed effects.  
Shown for reference.

```{r}
#| label: mod0
mod0 <- lmer(score ~ 1 + (1 | patient_id), data = dta_long, REML = TRUE)
summary(mod0)
```

- **Random effects**: `Std.Dev` of `(Intercept)` for `patient_id` is $\hat\sigma_{u_0}$ — how much patients vary in their baseline.
- **Residual**: $\hat\sigma$ — within-patient noise after accounting for patient identity.
- **Fixed effects**: the grand intercept $\hat\beta_0$.


::: {.callout-warning title="SPSS: random intercept model"}
After showing how to fit a random intercept model in SPSS,  
the challenge is to repeat it for the simulated data.

How would you interpret the value for the intercept, and its significance ?
:::

##### Model 1a — Add continuous time (visit)

For longitudinal data the interest typically is in the trajectories over time, the model should account for that.

```{r}
#| label: mod1
mod1 <- lmer(score ~ visit + (1 | patient_id), data = dta_long, REML = FALSE)
tidy_gt(mod1, effects = "fixed")
```


::: {.callout-warning title="SPSS: random intercept with visit effect"}
After showing how to add a slope within the random intercept model, 
the challenge is to repeat it for the simulated data.

How to interpret the value for the slope,  
and how does the interpretation for the intercept change ?
:::

##### Model 1b — Add categorical time (visit)

As a side note, when the trajectory can not be interpreted on a continuous scale, only between consecutive time points, time can be included as a categorical variable.

```{r}
#| label: mod1b
mod1b <- lmer(score ~ factor(visit) + (1 | patient_id), data = dta_long, REML = FALSE)
tidy_gt(mod1b, effects = "fixed")
```

::: {.callout-warning title="SPSS: random intercept with visit effect"}
After showing how to add set of slopes to capture changes over time within the random intercept model, 
the challenge is to repeat it for the simulated data.

How to interpret the values for the slopes,  
and how does the interpretation for the intercept change ?
:::

##### Model 2 — Add treatment

```{r}
#| label: mod2
mod2 <- lmer(score ~ visit + treatment + (1 | patient_id),
             data = dta_long, REML = FALSE)
tidy_gt(mod2, effects = "fixed")
```

::: {.callout-warning title="SPSS: add treatment effect"}
After showing how to add a treatment effect,  
the challenge is to repeat it for the simulated data.

How to interpret the treatment effect ?
:::

##### Model 3 — Add age (full fixed effects)

```{r}
#| label: mod3
mod3 <- lmer(score ~ visit + treatment + I(age-mean(age)) + (1 | patient_id),
             data = dta_long, REML = FALSE)
tidy_gt(mod3, effects = "fixed")
```

Centering, or scaling, can help interpretation, and for complex models make estimation more stable.


::: {.callout-warning title="SPSS: add continuous centered age effect"}
After showing how to center age and add it to the model,  
the challenge is to repeat it for the simulated data.

Hoe does the centering change the interpretation ?  
What if the age was standardized instead ?
:::

##### Model 4 — Add random slope for time

Allows each patient to have their own *rate of change* over time:

$$
\text{score}_{ij} = (\beta_0 + u_{0i}) + (\beta_1 + u_{1i})\,\text{visit}_{ij} + \beta_2\,\text{trt}_{i} + \beta_3\,\text{age}_{i} + \varepsilon_{ij}
$$

$$
\begin{pmatrix} u_{0i} \\ u_{1i} \end{pmatrix}
\sim \mathcal{N}\!\left(
\mathbf{0},\;
\begin{pmatrix} \sigma^2_{u_0} & \sigma_{u_0 u_1} \\ \sigma_{u_0 u_1} & \sigma^2_{u_1} \end{pmatrix}
\right)
$$

```{r}
#| label: mod4
mod4 <- lmer(score ~ visit + treatment + I(age-mean(age)) + (1 + visit | patient_id),
             data = dta_long, REML = FALSE)
tidy_gt(mod4, effects = "fixed")
```

Be awaret, not only should it be possible to fit a slope, there should be sufficient to estimate their variance well enough.


::: {.callout-warning title="SPSS: random slope"}
After showing how to add a random slope,   
the challenge is to repeat it for the simulated data.

How to interpret the estimated value for the random slope.
:::

### Model Comparison {data-toc-title="Model Comparison"}

Evaluate parts of a model by comparing the model with and without that part. Use for example the Likelihood Ratio Test (LRT) or an AIC.

With maximized log-likelihood \(\ell(\cdot)\) the LRT is

$$
\mathrm{LRT} = 2\left\{\ell(\text{complex model}) - \ell(\text{reduced model})\right\}.
$$

Important:  
to evaluate fixed-effects, use Maximum Likelihood (ML)  
and only use Restricted ML (REML) to evaluate random-effects.

#### Evaluation of fixed effects 

Models estimated with ML can be compared.

```{r}
#| label: anova-fixed
anova(mod1, mod2, mod3)
```

Note: to compare model 3 with model 2,  
the higher log-likelihood is of the more elaborate model (6 parameters)
the lower log-likelihood is of the less elaborate model (5 parameters)

evaluate the probability for a a higher value than 
the double the difference (-2*LL) on a $\chi^2$ distribution  
with degrees of freedom equal to the difference in number of parameters.


::: {.callout-warning title="SPSS: LRT for fixed effects"}
After showing how to perform an LRT for fixed effects,  
the challenge is to repeat it for the simulated data.

Can you compare model 3 and model 1 ? (p=.38)
:::

#### Evaluation of random effects

```{r}
#| label: anova-random

# Re-fit mod3 and mod4 with ML (already done above)
anova(mod3, mod4)
```

Parameters are at the boundary of the paramaterspace when testing whether a variance is 0, special care is required:

Testing a random intercept against no random effect, we compare

$$
H_0: \sigma^2_{b_0} = 0
\qquad \text{vs} \qquad
H_1: \sigma^2_{b_0} > 0.
$$

With \(\sigma^2_{b_0}=0\) on the boundary, the asymptotic null distribution of the LRT is
$$
\mathrm{LRT} \;\xrightarrow{d}\; \tfrac{1}{2}\chi^2_0 + \tfrac{1}{2}\chi^2_1.
$$

Equivalently, the p-value is
$$
p = \tfrac{1}{2} \Pr\left(\chi^2_1 \ge \mathrm{LRT}\right).
$$

For testing whether a random slope is needed assuming the random intercept is required, acknowledging that with a random slope there could also be random covariance:

$$
H_0: \sigma^2_{b_1} = 0
\qquad \text{vs} \qquad
H_1: \sigma^2_{b_1} > 0.
$$

The asymptotic null distribution of the likelihood ratio statistic:

$$
\mathrm{LRT} \;\xrightarrow{d}\; \tfrac{1}{2}\chi^2_1 + \tfrac{1}{2}\chi^2_2,
$$

For complex models a model-specific mixture or a parametric bootstrap is typically preferable.


::: {.callout-warning title="SPSS: LRT for random effects"}
After showing how to do a LRT for random effedcts,  
the challenge is to repeat it for the simulated data.

Is a random slope required ?
:::

#### AIC / BIC summary table

An alternative method for model evaluation and building are information criteria. AIC is standard, BIC is interesting to push more towards simpler models.

```{r}
#| label: model-comparison-table

model_list <- list(
  "M0: intercept only"       = mod0,
  "M1: + visit"              = mod1,
  "M2: + treatment"          = mod2,
  "M3: + age (full FE)"      = mod3,
  "M3*: - treatment"          = update(mod3,.~.-treatment),
  "M4: + random slope"       = mod4
)

map_dfr(model_list, glance, .id = "model") |>
  select(model, AIC, BIC, logLik, df.residual) |>
  mutate(across(where(is.numeric), \(x) round(x, 1))) |>
  gt() |>
  tab_style(
    style = cell_fill(color = "#d4edda"),
    locations = cells_body(rows = AIC == min(AIC))
  )
```


::: {.callout-warning title="SPSS: AIC and BIC to compare models"}
After showing how to extract the AIC and BIC,  
the challenge is to repeat it for the simulated data.

Do the conclusions confirm the LRT based conclusions ?
:::

Following the data the model of choice would be model 3 minus the treatment. Because we assume treatment is of primary concern in current study, it is nevertheless retained.

### Fixed and Random Effect Estimates {data-toc-title="Estimates"}

The model is selected (**mod3** — random intercept with full fixed effects) and re-fitted with REML for unbiased variance estimates.

```{r}
#| label: final-model

mod_final <- lmer(score ~ visit + treatment + scale(age) + (1 | patient_id),
                  data = dta_long, REML = TRUE)
```

For the resulting model, the estimates are considered.


::: {.callout-warning title="SPSS: Refit finalized model with REML"}
After showing the refitting of a final model,  
the challenge is to repeat it for the simulated data.

Does it matter, REML or ML ?
:::

#### Fixed effects

```{r}
#| label: tab-fixed-effects
tidy_gt(mod_final, effects = "fixed", conf.int = TRUE)
```

::: {.callout-warning title="SPSS: Test for the fixed effect"}
After showing how to test using a LRT for the fixed effect for treatment,  
the challenge is to repeat it for the simulated data.

How to interpret the estimated treatment effect ?
:::

#### Variance components

```{r}
#| label: tab-random-effects
tidy_gt(mod_final, effects = "ran_pars")
```


::: {.callout-warning title="SPSS: Test for the (co-) variances"}
After showing how to test using a LRT for the random parameters,  
the challenge is to repeat it for the simulated data,  
focus on testing whether a random slope should be added.

How to interpret the LRT test p-value ?
:::


#### Intraclass Correlation Coefficient (ICC)

The ICC quantifies the proportion of the **total variance**  attributable to between-patient differences (in random intercepts models):

Remember:

$$
ICC = \frac{\sigma^2_{u_0}}{\sigma^2_{u_0} + \sigma^2}
$$

```{r}
#| label: icc

vc <- as.data.frame(VarCorr(mod_final))
sigma2_u0  <- vc$vcov[vc$grp == "patient_id"]
sigma2_eps <- vc$vcov[vc$grp == "Residual"]
icc <- sigma2_u0 / (sigma2_u0 + sigma2_eps)

tibble(
  sigma2_u0  = round(sigma2_u0, 3),
  sigma2_eps = round(sigma2_eps, 3),
  ICC        = round(icc, 3)
) |> gt()
```


::: {.callout-warning title="SPSS: Extract the ICC"}
After showing how to extract the ICC,  
the challenge is to repeat it for the simulated data.
:::

As the ICC gets higher, the need for accounting correlation within the model, making use of a mixed model, increases.

### Prediction

After estimation, using the estimates, a prediction can be made for 

- patients present (in-bag)
- new / fictional patients (out-of-bag)

Predictions depend on what is in the model,  
with various building blocks in the model decisions have to be made.

Predictions often are a good tool to understand the model,  
to see what it implies.

#### Population-Level Predictions {data-toc-title="Population Predictions"}

Population-level predictions use only the **fixed effects** (marginal predictions):

$$
\hat{y}_{ij}^{\,\text{pop}} = \hat\beta_0 + \hat\beta_1\,\text{visit}_{ij} + \hat\beta_2\,\text{trt}_{i} + \hat\beta_3\,\text{age}_{i}
$$

```{r}
#| label: fig-population-predictions
#| fig-cap: "Population-level predicted trajectories by treatment"
#| fig-height: 4.5

new_pop <- expand_grid(
  visit     = 1:5,
  treatment = c("C", "T"),
  age       = mean(dta_long$age)   # at average age
)

new_pop <- new_pop |>
  mutate(pred_pop = predict(mod_final, newdata = new_pop, re.form = NA))

ggplot(new_pop, aes(x = visit, y = pred_pop, colour = treatment, group = treatment)) +
  geom_line(linewidth = 1.6) +
  geom_point(size = 3) +
  scale_colour_manual(values = c(C = col_patient, T = col_accent)) +
  labs(
    x = "Visit", y = "Predicted SQUARE test score",
    title = "Population-level predicted trajectories",
    subtitle = "At average age; random effects set to zero"
  ) +
  theme_bw(base_size = 13)
```

This prediction assumes an average age patient (our decision), for both Control and Treatment conditions at each time point. This plot can be generated for different ages.


::: {.callout-warning title="SPSS: Population level prediction"}
After showing the prediction of the general average accounting for within patient correlations,  
the challenge is to repeat it for the simulated data.
:::

#### Subject-Specific Predictions {data-toc-title="Subject Predictions"}

Subject-specific predictions use what is known after estimation. 

Predictions depend on the fixed effects (conditional average),  
and on the random effects (group specific deviations) following each patient's random intercept $\hat u_{0i}$:

$$
\hat{y}_{ij}^{\,\text{subj}} = \hat\beta_0 + \hat\beta_1\,\text{visit}_{ij} + \hat\beta_2\,\text{trt}_{i} + \hat\beta_3\,\text{age}_{i} + \hat u_{0i}
$$

```{r}
#| label: fig-subject-predictions
#| fig-cap: "Subject-specific predicted vs. observed trajectories (first 12 patients)"
#| fig-height: 6

dta_pred <- dta_long |>
  mutate(
    pred_pop  = predict(mod_final, re.form = NA),
    pred_subj = predict(mod_final)
  )

dta_pred |>
  filter(patient_id <= 12) |>
  ggplot(aes(x = visit)) +
  geom_point(aes(y = score), colour = "grey40", size = 2) +
  geom_line(aes(y = pred_subj, colour = treatment), linewidth = 1.1) +
  geom_line(aes(y = pred_pop, colour = treatment),
            linewidth = 0.7, linetype = "dashed", alpha = 0.6) +
  scale_colour_manual(values = c(C = col_patient, T = col_accent)) +
  facet_wrap(~patient_id, ncol = 4, labeller = label_both) +
  labs(
    x = "Visit", y = "Score",
    title = "Subject-specific (solid) vs. population (dashed) predictions",
    subtitle = "Points = observed scores"
  ) +
  theme_bw(base_size = 11) +
  theme(legend.position = "top")
```

::: {.callout-warning title="SPSS: Patient level prediction"}
After showing the prediction of a particular patients' outcome,  
the challenge is to repeat it for the simulated data.
:::

#### Predicting a New Patient {data-toc-title="New Patient"}

For a **new patient** not in the dataset, we cannot estimate their $u_{0i}$ — we set it to zero (population average). This is the **marginal prediction**.

```{r}
#| label: new-patient-prediction

new_patient <- tibble(
  visit     = 1:5,
  treatment = "T",
  age       = 40,
  patient_id = NA_integer_    # new patient
)

new_patient <- new_patient |>
  mutate(
    pred = predict(mod_final, newdata = new_patient, re.form = NA,
                   allow.new.levels = TRUE)
  )

new_patient |>
  select(visit, treatment, age, pred) |>
  mutate(pred = round(pred, 2)) |>
  gt() |>
  tab_header(
    title    = "Predicted scores for a new 40-year-old patient, treatment B",
    subtitle = "Random intercept set to 0 (population average)"
  )
```

::: {.callout-warning title="SPSS: New patient level prediction"}
After showing the prediction of a new patients' outcome,  
the challenge is to repeat it for the simulated data.
:::

### Diagnostics

The results from estimating and testing are conditional on the model and its appropriateness.  

A model is virtually never fully correct, an evaluation should be performed to identify possible causes of bias.

Good diagnostics for LMMs check:

1. Linearity and homoscedasticity of **residuals**
2. Normality of **residuals**
3. Normality of **random effects** (BLUPs)
4. Influential observations

```{r}
#| label: diagnostic-setup

res_df <- dta_long |>
  mutate(
    fitted_val = fitted(mod_final),
    resid_val  = resid(mod_final),
    std_resid  = resid_val / sigma(mod_final)
  )

blups <- ranef(mod_final)$patient_id |>
  rownames_to_column("patient_id") |>
  rename(u0_hat = `(Intercept)`)
```

#### Residuals vs. Fitted

Residuals should be **randomly scattered** around zero with **constant spread**. 

- systematic curve suggests a missing non-linear term
- funnel shape suggests heteroscedasticity.

```{r}
#| label: fig-resid-fitted
#| fig-cap: "Residuals vs. fitted values"
#| fig-height: 4

res_df |>
  ggplot(aes(x = fitted_val, y = resid_val)) +
  geom_point(alpha = 0.6, colour = col_patient) +
  geom_hline(yintercept = 0, linetype = "dashed", colour = col_grey) +
  geom_smooth(se = FALSE, colour = col_pop, linewidth = 1) +
  labs(x = "Fitted values", y = "Residuals",
       title = "Residuals vs. Fitted") +
  theme_bw(base_size = 13)
```

::: {.callout-warning title="SPSS: Residual Analysis over fitted values"}
After showing the residuals as a function of the fitted values,  
the challenge is to repeat it for the simulated data.
:::

#### Q–Q Plot of Residuals

Residuals should ideally show a **Normal** distribution. 

- normality follows from the addition of independent random perturbations
- normality follows from the central limit theorem

```{r}
#| label: fig-qq-resid
#| fig-cap: "Q–Q plot of standardised residuals"
#| fig-height: 4

res_df |>
  ggplot(aes(sample = std_resid)) +
  stat_qq(alpha = 0.6, colour = col_patient) +
  stat_qq_line(colour = col_pop, linewidth = 1) +
  labs(title = "Normal Q–Q: Standardised Residuals",
       x = "Theoretical Quantiles", y = "Sample Quantiles") +
  theme_bw(base_size = 13)
```

::: {.callout-warning title="SPSS: Residual Analysis, Normality"}
After showing the normality of residuals using a qqplot,  
the challenge is to repeat it for the simulated data.
:::

#### Normality of Random Effects (BLUPs)

Random effects typically are assumed following samples from a normal distribution.

- normality of groupspecific effects follows from the addition of independent random perturbations
- normality follows from the central limit theorem

```{r}
#| label: fig-qq-blups
#| fig-cap: "Q–Q plot of estimated random intercepts"
#| fig-height: 4

blups |>
  ggplot(aes(sample = u0_hat)) +
  stat_qq(alpha = 0.7, colour = col_accent) +
  stat_qq_line(colour = col_pop, linewidth = 1) +
  labs(title = "Normal Q–Q: Random Intercepts (BLUPs)",
       x = "Theoretical Quantiles", y = "Sample Quantiles") +
  theme_bw(base_size = 13)
```


::: {.callout-warning title="SPSS: Group-specific residual Analysis, Normality"}
After showing the normality of residuals using a qqplot,  
the challenge is to repeat it for the simulated data.
:::

#### Scale–Location Plot

The square root of absolute valued residuals makes it easier to discover heteroscedasticity, here as a function of the fitted value again. The smoother, a loess line, helps diagnostics.

```{r}
#| label: fig-scale-location
#| fig-cap: "Scale–location plot (homoscedasticity check)"
#| fig-height: 4

res_df |>
  mutate(sqrt_abs_resid = sqrt(abs(std_resid))) |>
  ggplot(aes(x = fitted_val, y = sqrt_abs_resid)) +
  geom_point(alpha = 0.6, colour = col_patient) +
  geom_smooth(se = FALSE, colour = col_pop, linewidth = 1) +
  labs(x = "Fitted values", y = expression(sqrt("|Standardised residuals|")),
       title = "Scale–Location") +
  theme_bw(base_size = 13)
```

::: {.callout-warning title="SPSS: Residual Analysis over fitted values"}
After showing the square rooted absolute valued residuals as a function of the fitted values,  
the challenge is to repeat it for the simulated data.
:::

#### BLUP Caterpillar Plot

Using the estimated heterogeneity, variance over patients, confidence intervals can be obtained around the random effects for each particular patient.  

```{r}
#| label: fig-caterpillar
#| fig-cap: "Caterpillar plot of random intercepts per patient"
#| fig-height: 5.5

sigma_u <- sqrt(as.data.frame(VarCorr(mod_final))$vcov[1])

blups |>
  arrange(u0_hat) |>
  mutate(
    patient_id = factor(patient_id, levels = patient_id),
    lower = u0_hat - 1.96 * sigma_u / sqrt(nrow(blups)),
    upper = u0_hat + 1.96 * sigma_u / sqrt(nrow(blups))
  ) |>
  ggplot(aes(x = patient_id, y = u0_hat)) +
  geom_hline(yintercept = 0, linetype = "dashed", colour = col_grey) +
  geom_errorbar(aes(ymin = lower, ymax = upper),
                width = 0.3, colour = col_grey, alpha = 0.7) +
  geom_point(size = 2.5, colour = col_patient) +
  labs(
    x = "Patient ID (sorted)", y = expression(hat(u)[0][i]),
    title = "Caterpillar plot: estimated random intercepts",
    subtitle = "Patients above zero tend to score higher than average"
  ) +
  theme_bw(base_size = 12) +
  theme(axis.text.x = element_text(angle = 90, hjust = 1, size = 7))
```

::: {.callout-warning title="SPSS: Residual Analysis over fitted values"}
After showing the extraction of confidence intervals for the random effects,  
the challenge is to repeat it for the simulated data.
:::
