## Borrowing Strength: Partial Pooling  {data-toc-title="Partial Pooling"}

Strategies for estimating each clusters' intercept:

1. **No pooling**: independent clusters, ignores information from other clusters
2. **Complete pooling**: grand mean, ignores group membership
3. **Partial pooling**: reconciles both, compromise

**Partial pooling**, the key mixed model characteristic

- cluster specific estimates
- more global estimates

Possible becasue clusters sampled from an underlying population

- other clusters are informative
- use dependency through common population of clusters

Partial pooling implies shrinkage:  
shrink each group estimate toward the grand mean  
results in random effects $\hat{u}_{0j}$ or **Best Linear Unbiased Predictor (BLUP)**:

$$
\hat{u}_{0j} = \underbrace{\frac{n_j/\sigma^2}{n_j/\sigma^2 + 1/\tau_{00}}}_{\text{reliability weight}\ \lambda_j} \cdot \left(\bar{y}_{\cdot j} - \hat{\gamma}_{00}\right)
$$

The weight defines the percentage with which cluster specific deviation from the global average shrink

- $\lambda_j \to 1$ with increasing cluster sizes and stronger within cluster correlations (trust the raw data)
- $\lambda_j \to 0$ witch decreasing cluster sizes and weaker within cluster correlations (shrink toward the overall mean).

```{r}
#| label: fig-partial-pooling-try
#| fig-cap: "Partial pooling. blue = no-pooling, green = LMM BLUPs"
#| fig-height: 5
#| fig-width: 10


set.seed(123)
# combined 7, 8, 9, to get 16 at the bottom
# split 6 in sizes 4, 4, and 8
# randomly assign in 1, 2, 3 and 4


dta_grouped_arrange_9 <- dta_grouped
dtb1 <- dta_grouped_arrange_9 |> filter(group %in% c(1:4)) |> mutate(group1=sample(group))
dtb2 <- dta_grouped_arrange_9 |> filter(group %in% c(5)) |> mutate(group1=factor(sample(c(rep(5,4),rep(6,4),rep(8,8)))))
dtb3 <- dta_grouped_arrange_9 |> filter(group %in% c(7)) |> mutate(group1=factor(7))
dtb4 <- dta_grouped_arrange_9 |> filter(group %in% c(7,8,9)) |> mutate(group1=factor(9))
dta_grouped_arrange_9 <- bind_rows(dtb1,dtb2,dtb3,dtb4)


lmer_int_9x <- lmer(score ~ age + (1 | group1), data = dta_grouped_arrange_9)
grand_mu <- fixef(lmer_int_9x)[["(Intercept)"]]

raw_means <- do.call(rbind, lapply(split(dta_grouped_arrange_9 |> group_by(group1) |> mutate(n_obs=n()), dta_grouped_arrange_9$group1), function(d) {
  data.frame(
    group = as.character(d$group1[1]),
    n_obs = d$n_obs[1],
    raw_mean = mean(d$score) - fixef(lmer_int_9x)[[2]] * mean(d$age)
  )
}))

re_vals <- ranef(lmer_int_9x)$group
re_df   <- data.frame(group = rownames(re_vals), u0 = re_vals[, 1])
re_df$blup <- grand_mu + re_df$u0

pp_data <- merge(raw_means, re_df, by = "group")
pp_data$gord <- factor(pp_data$group, levels = as.character(1:nrow(pp_data)))

ggplot(pp_data, aes(x = gord)) +
  geom_hline(yintercept = grand_mu, linetype = "dashed",
             color = "grey50", linewidth = 1) +
  geom_segment(aes(xend = gord, y = raw_mean, yend = blup),
               arrow = arrow(length = unit(0.18, "cm"), type = "closed"),
               color = "#e05c2a", linewidth = 0.7) +
  geom_point(aes(y = raw_mean, size = n_obs), color = "#2c6fad", alpha = 0.85) +
  geom_point(aes(y = blup), shape = 18, size = 3.5, color = "#3aab6d") +
  scale_size_continuous(range = c(2, 8), breaks = c(3, 8, 15, 30, 50)) +
  annotate("text", x = 12, y = grand_mu + 1.5,
           label = "Grand mean", size = 3, color = "grey50", hjust = 1) +
  labs(title = "Partial Pooling: Shrinkage toward the Grand Mean",
       subtitle = "Blue = no-pooling estimate  |  Green = LMM BLUP  |  Arrow = shrinkage",
       x = "Group", y = "Estimated intercept", size = "Group n") +
  theme(legend.position = "right")
```

Note: modeling the correlation to shrink estimates allows for cluster specific effects not explicitely represented

Example: matching is not a reason to turn to mixed models,  
matching uses explicitly available information that should included in the model as such

## Within-Cluster Correlation and Effective Sample Size {data-toc-title="Sample Size"}

Consider two types of predictors:

1. **Between-cluster predictor** $Z_j$ (e.g. hospital type)
   - Only varies between clusters, constant within a cluster.
2. **Within-cluster predictor** $x_{ij}$ (e.g. treatment dose)
   - Varies within clusters.

For a balanced design with $J$ clusters and $m$ observations per cluster:

- Variance of the estimator for the **between-cluster** effect is approximately proportional to

$$
\text{Var}(\hat{\gamma}_{01}) \propto \frac{\tau_{00} + \sigma^2/m}{J}.
$$

- Variance of the estimator for the **within-cluster** effect is approximately proportional to

$$
\text{Var}(\hat{\gamma}_{10}) \propto \frac{\sigma^2}{J \cdot m}.
$$

As $\rho$ increases (i.e. $\tau_{00}$ grows relative to $\sigma^2$):

- $\tau_{00} + \sigma^2/m$ becomes larger $\Rightarrow$ **between-cluster** effects become harder to detect.
- $\sigma^2$ becomes smaller relative to the total variance $\Rightarrow$ **within-cluster** effects become easier to detect.

**Stronger within-cluster correlation** makes 

- **between-cluster differences** harder to detect.
- **within-cluster differences** easier to detect.

Note: cluster randomization focuses on between cluster effects, dependent more on the number of clusters, less on the number of observations.

To compare treatments,  
**randomizing within each cluster is usually far more efficient**

```{r}
save.image('zenzi.RData')
```

