---
title: "Mixed Models: a Primer"
subtitle: "About how to analyze clustered data"
date: today
published-title: "Rendered"
---

```{r}
#| label: packages
pckgs <- c('tidyverse', 'scales', 'gt', 'broom','lme4','nlme','knitr','patchwork','broom.mixed','lmerTest','emmeans')
donotshow <- purrr::map(pckgs, library, character.only = TRUE)
select <- dplyr::select

options(pillar.sigfig = 4, pillar.neg = "zigzag")
```

```{r}
#| label: my_functions
tidy_gt <- function(model, type = c("summary", "anova"), decimals = 3, ...) {
  type <- match.arg(type)
  
  if (type == "summary") {
    out <- tidy(model, ...)
    gt(out) |>
      fmt_number(columns = c(estimate, std.error, statistic), decimals = decimals) |>
      fmt_scientific(columns = p.value) |>
      cols_label(
        term = "Term",
        estimate = "Estimate", 
        std.error = "Std. Error",
        statistic = "t value",
        p.value = "p value"
      ) |>
      tab_options(table.font.size = 14)
      
  } else {
    out <- tidy(anova(model), ...)
    gt(out) |>
      fmt_number(
        columns = c(df, sumsq, meansq, statistic), 
        decimals = decimals
      ) |>
      fmt_scientific(columns = p.value) |>
      cols_label(
        term = "Term",
        df = "df",
        sumsq = "Sum Sq",
        meansq = "Mean Sq",
        statistic = "F value",
        p.value = "p value"
      ) |>
      tab_options(table.font.size = 14)
  }
}
glance_gt <- function(model, decimals = 3, ...) {
  glance(model, ...) |> 
    gt() |> 
    fmt_number(decimals = decimals) |> 
    cols_label(
      r.squared = html("R²"),
      adj.r.squared = html("adj. R²"),
      sigma = "σ",
      AIC = "AIC", BIC = "BIC",
      logLik = "logLik"
    ) |> 
    tab_options(table.font.size = 14)
}
tidy_mixed_gt <- function(model, ...) {
  broom.mixed::tidy(model, ...) |>
    mutate(across(where(is.numeric), \(x) round(x, 4))) |>
    gt() |>
    tab_options(table.font.size = 13)
}

glance_mixed_gt <- function(model) {
  broom.mixed::glance(model) |>
    mutate(across(where(is.numeric), \(x) round(x, 3))) |>
    gt() |>
    tab_options(table.font.size = 13)
}

```

```{r}
#| label: default_theme
theme_lmm <- function(base_size = 11) {
  theme_minimal(base_size = base_size) +
    theme(
      plot.title    = element_text(size = 16, face = "bold", margin = margin(b = 6)),
      plot.subtitle = element_text(size = 14, color = "#555", margin = margin(b = 10)),
      axis.title    = element_text(size = 14),
      axis.text.x   = element_text(size = 10),
      axis.text.y   = element_text(size = 10),
      axis.line     = element_line(color = "black"),
      panel.grid.minor = element_blank(),
      panel.grid.major = element_line(color = "#efefef"),
      legend.position  = "bottom",
      legend.title     = element_text(size = 14),
      legend.text      = element_text(size = 14),
      strip.text       = element_text(face = "bold", size = 12)
    )
}
theme_set(theme_lmm())
```

::: {.callout-note appearance="simple"}
**Vitality turns Mixed**

- Aim: develop intuition for mixed models  

&nbsp;

- Focus on
    - equations (building blocks)
    - visualizations
:::

```{r}
load('zenzi.RData')
```

# Context {.unlisted .unnumbered} 

Long History

- 1950: Henderson equations define mixed model
- 1982: Laird and Ware model allows modeling growth trajectories
- 1990: SAS PROC MIXED makes it possible for us, now in all major software

Added Value

- MANOVA made flexible and efficient
- Extends regression; modeling dependencies


{{< include ws-mixed-0-prep.qmd >}}
{{< include ws-mixed-1-regress.qmd >}}
{{< include ws-mixed-2-cluster.qmd >}}
{{< include ws-mixed-3-model.qmd >}}
{{< include ws-mixed-4-pooling.qmd >}}
{{< include ws-mixed-5-slope.qmd >}}
{{< include ws-mixed-6-higher.qmd >}}
{{< include ws-mixed-7-longitudinal.qmd >}}
{{< include ws-mixed-9-finalize.qmd >}}

<!--


-->



```{r}
save.image('zenzi.RData')
```