# Mixed Model

Mixed models (a.k.a Multilevel Model)

- extend regression models,
- is modeling variance
- turning it into a conditional mean
- by estimating relations
- by minimizing squared error.
- **while accomodating dependencies**
- **through estimating variances (heterogeneity)**

## Random Intercept Model  {data-toc-title="Random Intercept"}

Let $y_{ij}$ denote the outcome for individual $i$ in cluster $j$.

::: {.equation-box}
**Level 1** (case within cluster):
$$
y_{ij} = \underline{\beta_{0j}} + \beta_1 x_{ij} + \varepsilon_{ij}, \qquad \varepsilon_{ij} \sim \mathcal{N}(0,\ \sigma^2)
$$

**Level 2** (cluster-level model for the intercept):
$$
\underline{\beta_{0j}} = \gamma_{00} + u_{0j}, \qquad u_{0j} \sim \mathcal{N}(0,\ \tau_{00})
$$

**Combined (reduced form)**:
$$
y_{ij} = \underbrace{\gamma_{00} + \beta_1 x_{ij}}_{\text{fixed part}} + \underbrace{u_{0j} + \varepsilon_{ij}}_{\text{random part}}
$$
:::

The random part implies a variance decomposed into random intercept variance and residual variance:

$$
\text{Var}(y_{ij}) = \tau_{00} + \sigma^2, \qquad \text{Cov}(y_{ij},\ y_{i'j}) = \tau_{00} \quad (i \neq i', \text{ same group } j)
$$

The relative proportion of heterogeity for this simple model is the intra class correlation (ICC):

$\text{ICC} = \tau_{00} / (\tau_{00} + \sigma^2)$.

## Why a Random Intercept Model  {data-toc-title="Advantage"}

Mixed models treat cluster effects as **random draws from a distribution**.  

Instead of estimating J-1 fixed effects + intercept

$$
\beta_{0}, \beta_{1}, ..., \beta_{J-1}
$$

random effects are estimated as

$$
\gamma_{00} + u_{0j} \qquad with \qquad u_{0j} \sim \mathcal{N}(0,\ \tau_{00})
$$

resulting in only 2 parameters to estimate: grand mean $\gamma_{00}$ and variance $\tau_{00}$

Notice the similarity with the residuals:

$$
\varepsilon_i \overset{\text{i.i.d.}}{\sim} \mathcal{N}(0, \sigma^2).
$$

A single variance parameter captures cluster heterogeneity

With a growing number of clusters

- no increase in number of parameters to estimate
- a better estimate of the variance parameter

In this particular case:

```{r}
#| label: lmer_random_intercept
lmer_int_9 <- lmer(score ~ age + (1 | group), data = dta_grouped)
```

```{r}
#| label: fig-mixed_grouped_fitted
cont9_stats <- dta_grouped |>
  summarise(
    mean_score = mean(score),
    sd_score   = sd(score),
    n          = n(),
    se_mean    = sd(score) / sqrt(n()),
    .groups    = "drop"
  )

pltdta_ri_9_high <- augment(lmer_int_9, data = dta_grouped) |> select(-resid) |> rename(fitted = .fitted, resid = .resid)

pred_data <- expand_grid(
  age = seq(min(dta_grouped$age), max(dta_grouped$age), length.out = 20),
  group = factor(1:9)
)

pred_data$fitted <- predict(lmer_int_9, newdata = pred_data, re.form = NULL)
pred_data$group_gray <- factor(pred_data$group, levels = unique(pred_data$group)[order(c(1,5,9,2,6,3,7,4,8))])

```

```{r}
#| label: fig-mixed_age
#| fig-cap: "Mixed Model"
#| fig-height: 5.5
#| fig-width: 5
plt_mixed_similarity_high <- ggplot(pltdta_ri_9_high, aes(x = age, y = score)) +
  geom_point(aes(color = factor(group)), size = 2.5, alpha = 0.85) +
  scale_color_manual(values = group_grays_9, guide = "none") + 
  geom_segment(
    data = pltdta_ri_9_high,
    aes(x = age, xend = age, y = score, yend = fitted),
    color = "gray60",
    alpha = 0.3,
    linewidth = 0.8
  ) +
  geom_line(
    data = pred_data,
    aes(y = fitted, color = group_gray),
    linewidth = 1.2,
    show.legend = FALSE
  ) +
  annotate(
    "text",
    x = mean(pltdta_ri_9_high$age),
    y = max(pltdta_ri_9_high$score) + 0.5,
    label = paste0(
      "score ≈ ",
      round(fixef(lmer_int_9)[1], 2),
      " + ",
      round(fixef(lmer_int_9)[["age"]], 3),
      " × age + random effect"
    ),
    size = 4.5
  ) +
  labs(
    x = "Age (continuous predictor)",
    y = "Score (outcome)",
    title = "Mixed Model"
  ) +
  coord_cartesian(ylim = c(10, 20))
```

```{r}
#| label: fig-fixed_age
#| fig-cap: "Fixed Model"
#| fig-height: 5.5
#| fig-width: 5
plt_fixed_similarity_high <- plt_similarity_high +
  labs(
    x = "Age (continuous predictor)",
    y = "Score (outcome)",
    title = "Fixed model"
  )
```

```{r}
#| label: fig-grouping-fixed-random
#| fig-cap: "Random vs. Fixed: in this particular case almost identical"
#| fig-height: 5.5
#| fig-width: 12
wrap_plots(plt_mixed_similarity_high, plt_fixed_similarity_high, nrow=1)
```

```{r}
#| label: get_lm_lmer
# maybe change later to tidy_gt and tidy_mixed_gt
t_lm  <- tidy(lm_group_main_9)  |> select(term, estimate, std.error, statistic) |> mutate(across(c(estimate, std.error, statistic), ~ round(.x, 3)))
m_lm <- glance(lm_group_main_9) |> select(sigma) |> mutate(sigma=round(sigma,3))
t_lmer <- tidy(lmer_int_9)   |> select(term, estimate, std.error, df, statistic) |> mutate(across(c(estimate, std.error, statistic, df), ~ round(.x, 3)))
```

:::: {.columns}


::: {.column width="50%"}

Mixed  
`in R:`  
`lmer(score ~ age + (1 | group),`  
`data = ..)`

```{r}
#| label: tab-gt_lmer
gt(t_lmer)
```

Cluster specific estimates extracted:

```{r}
#| label: tab-gt_lmer_ranef
gt(ranef(lmer_int_9)$group)
```

```{r}
#| label: tab-icc_manual
vc <- as.data.frame(VarCorr(lmer_int_9))
tau00_hat  <- vc$vcov[vc$grp == "group"]
sigma2_hat <- vc$vcov[vc$grp == "Residual"]
icc_val    <- tau00_hat / (tau00_hat + sigma2_hat)
icc_vals <- c('tau_00'=tau00_hat, 'sigma2'=sigma2_hat, 'ICC'=icc_val)
icc_df <- data.frame(icc_vals) %>%
  rownames_to_column("metric")
gt(icc_df)
```

:::

::: {.column width="5%"}

:::

::: {.column width="45%"}

Fixed  
`in R`  
`lm(score ~ age + group,`  
`data = ..)`

```{r}
#| label: tab-gt_lm
gt(t_lm)
```


Fixed, residuals

```{r}
#| label: tab-gt_lm_anova
gt(m_lm)
```

:::

::::

## Level specific slopes {data-toc-title="Level 2 Variables"}

Different levels can have different predictors of the observations (at level 1).

- example: test results explained by patient (level 1) hospital (level 2) variables
  - e.g., older patients (age) result in higher scores
  - e.g., university hospitals (type) result in higher scores

Note: Variables can only explain variability at the lower or equal level.  

Let $y_{ij}$ denote the outcome for individual $i$ in cluster $j$.

::: {.equation-box}
**Level 1** (case within cluster):
$$
y_{ij} = \beta_{0j} + \beta_{1} x_{ij} + \varepsilon_{ij}, \qquad \varepsilon_{ij} \sim \mathcal{N}(0,\ \sigma^2)
$$

**Level 2** (cluster-level model for the intercept, with a Level‑2 predictor $z_{j}$):
$$
\beta_{0j} = \gamma_{00} + \underline{\gamma_{01} z_{j}} + u_{0j}, \qquad u_{0j} \sim \mathcal{N}(0,\ \tau_{00})
$$

**Combined (reduced form)**:
$$
y_{ij} = \underbrace{\gamma_{00} + \beta_{1} x_{ij} + \gamma_{01} z_{j}}_{\text{fixed part}} + \underbrace{u_{0j} + \varepsilon_{ij}}_{\text{random part}}
$$
:::

The same variable can sometimes be split up and defined at different levels.
Let $y_{ij}$ denote the outcome for individual $i$ in cluster $j$.

::: {.equation-box}
**Level 1** (case within cluster):
$$
y_{ij} = \beta_{0j} + \underline{\beta_{1W} x_{ijW}} + \varepsilon_{ij}, \qquad \varepsilon_{ij} \sim \mathcal{N}(0,\ \sigma^2)
$$

where  
$x_{ijW} = x_{ij} - \bar{x}_{j}$  (within-cluster centered $x$)  
$\bar{x}_{j} = \frac{1}{n_j}\sum_{i=1}^{n_j} x_{ij}$  (cluster mean of $x$)

**Level 2** (cluster-level model for the intercept, now including the cluster mean):
$$
\beta_{0j} = \gamma_{00} + \underline{\gamma_{01}\bar{x}_{j}} + u_{0j}, \qquad u_{0j} \sim \mathcal{N}(0,\ \tau_{00})
$$

**Combined (reduced form)**:
$$
y_{ij} = \underbrace{\gamma_{00} + \beta_{1W} x_{ijW} + \gamma_{01}\bar{x}_{j}}_{\text{fixed part}} + \underbrace{u_{0j} + \varepsilon_{ij}}_{\text{random part}}
$$
:::

A fixed slope, maybe age, or weight, ... split in  

- within cluster centered slope
- cluster specific average

One part focusses on within cluster relative position, other on cluster properties.

Example: blood pressure explained by body weight 

- patients with relatively higher weights in a hospital may have higher blood pressure
- hospitals may differ in average weight because of ... 

Without this split, it could be a blend.

Note: ecological fallacy.

```{r}
dta_grouped_center <- dta_grouped |> group_by(group) |> mutate(agem=mean(age), agec=age-agem)
```

```{r}
#| label: fig-within-between
#| fig-cap: "Within and Between Group Differences"
#| fig-height: 5.5
#| fig-width: 12

# Plot 1: Within-group part (Level 1) - agec (centered age)
plt_center_within <- ggplot(data = dta_grouped_center, aes(x = agec, y = score)) +
  geom_point(alpha = 0.3) +
  geom_line(aes(group = group), alpha = 0.5) +
  geom_smooth(aes(group = 1), method = "lm", se = FALSE, color = "blue", linewidth = 1) +  # pooled within-group slope
  labs(title = "Within-group (Level 1)", 
       x = "Age (centered within group)", 
       y = "Score") +
  theme_minimal()

# Plot 2: Between-group part (Level 2) - group means
dta_group_means <- dta_grouped_center |> 
  group_by(group) |> 
  summarise(score_mean = mean(score), 
            agem = mean(age),
            .groups = "drop") |> 
  mutate(group = factor(group),
         group_ordered = factor(group, levels = order(agem)))  # order by agem

plt_center_between <- ggplot(data = dta_group_means, aes(x = group_ordered, y = score_mean)) +
  geom_point(size = 3) +
  geom_line(aes(group = 1), alpha = 0.5) +
  geom_smooth(method = "lm", se = FALSE, color = "red", size = 1) +  # between-group slope
  labs(title = "Between-group (Level 2)", 
       x = "Group (ordered by mean age)", 
       y = "Mean Score") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

# Combine plots
plt_center_within + plt_center_between + plot_layout(nrow = 1)
```

```{r}
#| label: lmer_random_intercept_centered
lmer_int_global_9 <- lmer(score~age+(1|group),data=dta_grouped_center)
lmer_int_center_9 <- lmer(score~agem+agec+(1|group),data=dta_grouped_center)
```

Repeating just fitting a slope.

```{r}
tidy_mixed_gt(lmer_int_global_9)
glance_mixed_gt(lmer_int_global_9)
```

Splitting up over levels.


```{r}
tidy_mixed_gt(lmer_int_center_9)
glance_mixed_gt(lmer_int_center_9)
```

For the data, a general slope can be fitted, and alternatively, a within and a between slope.

```{r}
#| eval: false
lmer_int_slope_0_18 <- lmer(score~age+(1|group),data=dta_grouped_center)
lmer_int_slope_C_18 <- lmer(score~agem+agec+(1|group),data=dta_grouped_center)
```


```{r}
#| eval: false
dta_grouped_x2 <- dta_grouped_center
lmer_int_slope_0_18 <- lmer(score~age+(1|group),data=dta_grouped_x2)
lmer_int_slope_C_18 <- lmer(score~age_ind+age_group+(1|group),data=dta_grouped_x2 |> group_by(group) |> mutate(age_group=mean(age)) |> mutate(age_ind=age-age_group))
```

```{r}
#| eval: false
tidy_mixed_gt(lmer_int_slope_0_18)
glance_mixed_gt(lmer_int_slope_0_18)
```


```{r}
#| eval: false
tidy_mixed_gt(lmer_int_slope_C_18)
glance_mixed_gt(lmer_int_slope_C_18)
```

Relative age matters within clusters,  
Cluster average ages does not matter.

```{r}
save.image('zenzi.RData')
```