---
title: "Mixed-Effects Model Analysis of Orthodont Data"
author: "Statistical Analysis Example"
date: "`r format Sys.Date() %Y-%m-%d`"
format:
  html:
    theme: cosmi
    toc: true
---

```{r}
# Load required packages
library(nlme)      # For Orthodont data and lme()
library(lme4)      # For alternative lmer() implementation
library(ggplot2)   # For visualization
library(dplyr)     # For data manipulation
```

## 1. Data Overview

The `Orthodont` dataset contains measurements of the distance between the pituitary and pterygomaxillary fissure (in mm) for 27 children, measured at 4 different ages (8, 10, 12, and 14 years). The data also includes sex information.

```{r}
# Load and examine the data
data(Orthodont)
cat("Dataset structure:\\n")
str(Orthodont)
cat("\\nNumber of observations:", nrow(Orthodont), "\\n")
cat("Number of subjects:", length(unique(Orthodont$Subject)), "\\n")
cat("Age range:", min(Orthodont$age), "-", max(Orthodont$age), "years\\n")
cat("Sex distribution:\\n")
table(Orthodont$Sex)
```

```{r}
# Visualize the data
ggplot(Orthodont, aes(x = age, y = distance, color = Sex, group = Subject)) +
  geom_point(size = 2) +
  geom_line(alpha = 0.5) +
  stat_summary(fun = mean, geom = "point", size = 4, color = "black") +
  stat_summary(fun = mean, geom = "line", size = 1.5, color = "black") +
  labs(title = "Orthodont Distance by Age and Sex",
       x = "Age (years)",
       y = "Distance (mm)") +
  theme_minimal()
```

## 2. Model Strategy

We will fit a **full random intercept and random slope model** with:

- **Fixed effects**: Age, Sex, and their interaction
- **Random effects**: Random intercept and random slope for age by Subject (correlated)
- **Correlation structure**: None (independent residuals within subjects)

This is a more comprehensive model than the simple random intercept model.

## 3. Full Mixed Model using `nlme::lme()`

The model formula is:
$$Y_{ij} = \beta_0 + \beta_1 \text{Age}_{ij} + \beta_2 \text{Sex}_{i} + \beta_3 \text{Age}_{ij} \times \text{Sex}_{i} + b_{0i} + b_{1i} \text{Age}_{ij} + \epsilon_{ij}$$

where $b_{0i}$ (random intercept) and $b_{1i}$ (random slope) are correlated random effects for subject $i$.

```{r}
# Full mixed model with random intercept AND random slope (correlated)
fm_full_nlme <- lme(
  distance ~ age * Sex,              # Fixed effects: age, sex, interaction
  random = pdSymm(Subject ~ age),    # Random: correlated intercept & slope for age
  data = Orthodont,
  method = "REML"                    # Restricted Maximum Likelihood
)

cat("Model Summary (nlme::lme):\\n")
print(summary(fm_full_nlme))
```

```{r}
# Model diagnostics
cat("\\nResidual diagnostics:\\n")
par(mfrow = c(2, 2))
plot(fm_full_nlme)
```

## 4. Alternative Implementation using `lme4::lmer()`

The same model can be fitted using the `lme4` package:

```{r}
fm_full_lme4 <- lmer(
  distance ~ age * Sex + (1 + age | Subject),  # Same model structure
  data = Orthodont,
  REML = TRUE
)

cat("Model Summary (lme4::lmer):\\n")
print(summary(fm_full_lme4))
```

## 5. Model Comparison

Let's compare the full model with a simpler random-intercept-only model:

```{r}
# Random intercept only model
fm_restricted <- lme(
  distance ~ age * Sex,
  random = ~ 1 | Subject,           # Only random intercept
  data = Orthodont,
  method = "REML"
)

# Likelihood ratio test
anova(fm_restricted, fm_full_nlme)
```

The LR test evaluates whether the random slope significantly improves the model.

## 6. Random Effects Interpretation

```{r}
# Extract and examine random effects
cat("\\nRandom effects variance-covariance matrix:\\n")
print(var(fm_full_nlme))

cat("\\nRandom intercepts and slopes for first 5 subjects:\\n")
re <- random.effects(fm_full_nlme)
print(head(re, 5))
```

```{r}
# Plot subject-specific trajectories vs. population average
plot(fm_full_nlme, depth = 2)
```

## 7. Fixed Effects Interpretation

From the model summary, the fixed effects represent:

| Effect | Interpretation |
|--------|---------------|
| `(Intercept)` | Baseline distance at age 0 for Males |
| `age` | Change in distance per year for Males |
| `SexFemale` | Difference in baseline distance for Females vs. Males |
| `age:SexFemale` | Difference in age slope for Females vs. Males |

```{r}
# Calculate predicted values and confidence intervals
Orthodont$predicted <- predict(fm_full_nlme, level = 0)  # Population-level predictions
Orthodont$fitted <- predict(fm_full_nlme, level = 1)     # Subject-specific predictions

cat("\\nFixed effects with confidence intervals:\\n")
confint(fm_full_nlme)
```

## 8. Key Model Components Explained

### Fixed Effects (`distance ~ age * Sex`)
- **age**: Continuous predictor (years)
- **Sex**: Binary predictor (Male/Female)
- **age * Sex**: Interaction allowing different age slopes by sex

### Random Effects (`random = pdSymm(Subject ~ age)`)
- `pdSymm`: Specifies an unstructured (symmetric positive-definite) covariance matrix
- This allows the random intercept and slope to be **correlated**
- Alternative: `pdDiag()` would force them to be independent

### Estimation Method
- **REML**: Restricted Maximum Likelihood (preferred for final model)
- **ML**: Maximum Likelihood (needed for likelihood ratio tests between models)

## 9. Model Assessment

```{r}
# Check model assumptions
cat("Residual standard deviation:", fm_full_nlme$sigma, "\\n")
cat("Log-likelihood:", logLik(fm_full_nlme), "\\n")

# ICC (Intra-class Correlation Coefficient) approximation
var_components <- VarCorr(fm_full_nlme)
icc <- var_components / (var_components + var_components + fm_full_nlme$sigma^2)[1][2]
cat("\\nApproximate ICC:", round(icc, 3), "\\n")
```

## 10. Conclusion

This full mixed model:
- Accounts for **within-subject correlation** via random intercepts and slopes
- Allows **sex-specific growth trajectories** via the interaction term
- Estimates **correlation between intercept and slope** (children starting higher may grow faster/slower)
- Provides both **population-average** and **subject-specific** predictions

The model is appropriate for longitudinal data with repeated measurements per subject.

---

## Appendix: Complete R Code

```{r}
# Complete reproducible code block
library(nlme)
library(lme4)

data(Orthodont)

# Full model
fm_full <- lme(
  distance ~ age * Sex,
  random = pdSymm(Subject ~ age),
  data = Orthodont,
  method = "REML"
)

# Summary
summary(fm_full)

# Predictions
predict(fm_full, level = 0)  # Fixed effects only
predict(fm_full, level = 1)  # With random effects
```